---
title: "Flexible and Fast Modelling with McMasterPandemic -- Examples"
output: html_document
---

[![Live
Version](https://img.shields.io/static/v1.svg?label=Updated&message=`r Sys.Date()`&color=blue)](https://canmod.net/misc/flex_intro)

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, message=FALSE}
library(ggplot2)
library(McMasterPandemic)
library(dplyr)
library(tidyr)
library(lubridate)
```

## Hello World: Simulating an SIR Model

```{r}
state = c(S = 20000, I = 100, R = 0)
sir_model = (
  init_model(
    params = c(
      gamma = 0.06,
      beta = 0.15,
      N = sum(state)
    ),
    state = state,
    start_date = "2000-01-01",
    end_date = "2000-05-01",
    do_hazard = TRUE,
    do_make_state = FALSE
  )
  %>% add_rate("S", "I", ~ (1/N) * (beta) * (I))
  %>% add_rate("I", "R", ~ (gamma))
  %>% add_outflow()
  %>% add_tmb_indices
)
sir_model
```

```{r}
(sir_model
    %>% simulate_state_vector(add_dates = TRUE)
    %>% pivot_longer(!Date)
    %>% rename(state = value, epi_cat = name)
    %>% ggplot
     +  geom_line(aes(x = Date, y = state, colour = epi_cat))
  )
```

```{r}
(sir_model
    %>% simulate_changing_ratemat_elements(add_dates = TRUE)
    %>% rename(`force of infection` = S_to_I)
    %>% ggplot
     +  geom_line(aes(x = Date, y = `force of infection`))
  )
```

## Structure: Two-Strain SIR

```{r}

strains = c("wild", "variant")
state = c(
  S = 20000, 
  I_wild = 49, I_variant = 1,
  R_wild = 0,   R_variant = 0
)
two_strain_model = (
  init_model(
    params = c(
      gamma = 0.06,       
      beta_wild = 0.15,   
      beta_variant = 0.25,
      N = sum(state)
    ),
    state = state,
    start_date = "2000-01-01",
    end_date = "2000-05-01",
    do_hazard = TRUE,
    do_make_state = FALSE
  )
  %>% vec_rate(
    "S",
    "I" %_% strains, 
     vec("beta" %_% strains) * struc("1/N") * vec("I" %_% strains)
  )
  %>% rep_rate("I", "R", ~ (gamma))
  %>% add_outflow()
  %>% add_tmb_indices
)
```

```{r}
regex = "^(S|I|R)(_*)(.*)"
(two_strain_model
    %>% simulate_state_vector(add_dates = TRUE)
    %>% pivot_longer(!Date)
    %>% rename(state = value)
    %>% mutate(strain  = sub(pattern = regex, replacement = "\\3", name))
    %>% mutate(epi_cat = sub(pattern = regex, replacement = "\\1", name))
    %>% ggplot
     +  geom_line(aes(x = Date, y = state, colour = epi_cat, linetype = strain))
  )
```

```{r}
(two_strain_model
    %>% simulate_changing_ratemat_elements(add_dates = TRUE)
    %>% pivot_longer(starts_with("S_to_I"))
    %>% mutate(name = sub("S_to_I_", "", name))
    %>% rename(`force of infection` = value)
    %>% rename(strain = name)
    %>% ggplot
     +  geom_line(aes(x = Date, y = `force of infection`, colour = strain))
  )
```

## The SIRV model

```{r}
state = c(S = 20000, I = 100, R = 0, V = 0)
params = c(
      gamma = 0.06,
      beta = 0.15,
      v = 0, # initial vaccination rate
      N = sum(state)
    )

# roll out the vaccine by bumping the 
# vaccination rate twice
params_timevar = data.frame(
  Date = c("2020-02-01", "2020-03-01"), 
  Symbol = c("v", "v"),  
  Value = c(0.01, 0.1), 
  Type = c("abs", "abs"))

sirv_model = (
  init_model(
    params = params,
    state = state,
    params_timevar = params_timevar,
    start_date = "2020-01-01",
    end_date = "2020-05-01",
    do_hazard = TRUE,
    do_make_state = FALSE
  )
  %>% add_rate("S", "I", ~ (beta) * (1/N) * (I))
  %>% add_rate("I", "R", ~ (gamma))
  %>% add_rate("S", "V", ~  (v))
  %>% add_outflow()
  %>% add_tmb_indices
)

(sirv_model
    %>% simulate_state_vector(add_dates = TRUE)
    %>% pivot_longer(!Date)
    %>% rename(state = value, epi_cat = name)
    %>% ggplot
     +  geom_line(aes(x = Date, y = state, colour = epi_cat))
     +  geom_vline(aes(xintercept = as.Date(Date)), data = params_timevar, colour = 'lightgrey')
  )

```


## Variolation model

```{r}
state = c(
  S = 20000, 
  I_severe = 50, I_mild = 50,
  R_mild = 0,   R_severe = 0
)
params = c(
  nu = 0.0105, # birth rate
  mu = 0.0105, # death rate
  delta = 1/7, # waning immunity rate
  gamma_mild = 1/(3.3 + 9),  # recovery rate of mild cases
  gamma_severe = 1/(3.3 + 14), # recovery rate of severe cases
  beta_mild = 0.15,  # transmission rate of infections leading to mild cases
  beta_severe = 0.3, # transmission rate of infections leading to severe cases
  m = 0.6 # probability of developing mild illness
)

# model structure
all_states = names(state)
severity = c("mild", "severe")
beta_vec = vec("beta" %_% severity)
I_vec = vec("I" %_% severity)
foi = kronecker(vec("m", "1-m"), sum(beta_vec * I_vec * struc("1/N")))

variolation_model <- (
  init_model(
    params = params,
    state = state,
    start_date = "2020-01-01",
    end_date = "2020-04-01",
    do_hazard = FALSE,
    do_make_state = FALSE
  )
  %>% add_state_param_sum("N", any_var(state))
  
  # births and deaths 
  # (FIXME: this only works if nu = mu)
  %>% rep_rate(
    all_states,
    "S",
    ~ (nu)
  )
  
  # infection
  %>% vec_rate(
    "S",
    "I" %_% severity, 
     foi
  )
  
  # waning immunity
  %>% rep_rate(
    "R" %_% severity,
    "S",
    ~ (delta)
  )
  
  # recovery
  %>% vec_rate(
    "I" %_% severity,
    "R" %_% severity,
    vec("gamma" %_% severity)
  )
  
  %>% add_outflow()
  %>% add_tmb_indices
)

(variolation_model
    %>% simulate_state_vector(add_dates = TRUE)
    %>% pivot_longer(!Date)
    %>% rename(state = value, epi_cat = name)
    %>% ggplot
     +  geom_line(aes(x = Date, y = state, colour = epi_cat))
  )

```

## SEIRD Model

This is the [Mac Theo Bio Model](https://github.com/mac-theobio/mac-theobio.github.io/blob/master/seird.jpg).

```{r}
state = c(S = 20000, E = 50, I = 50, R = 0, D = 0, Null = 0)
params = c(
  # transmission rates of susceptibles from live and dead individuals
  beta_I = 5,
  beta_D = 2,
  
  # average times in various boxes
  T_E = 10,
  T_I = 14,
  T_D = 10,
  
  # probability of death given infection
  f = 0.5
)
seird_model = (
  init_model(
    params = params, 
    state = state,
    start_date = "2000-01-01",
    end_date = "2000-03-01",
    do_hazard = TRUE,
    do_make_state = FALSE
  )
  %>% add_state_param_sum("N", any_var(state))
  %>% add_rate("S", "E", ~ (beta_I) * (1/N) * (I) + (beta_D) * (1/N) * (D))
  %>% add_rate("E", "I", ~ (1/T_E))
  %>% add_rate("I", "R", ~ (1/T_I) * (1 - f))
  %>% add_rate("I", "D", ~ (1/T_I) * (f))
  %>% add_rate("D", "Null", ~ (1/T_D))
  %>% add_outflow()
  %>% add_tmb_indices
)
(seird_model
    %>% simulate_state_vector(add_dates = TRUE)
    %>% pivot_longer(!Date)
    %>% rename(state = value, epi_cat = name)
    %>% ggplot
     +  geom_line(aes(x = Date, y = state, colour = epi_cat))
  )
```

## Covid SEIR

The BC covid modelling group uses this [compartmental model](https://journals.plos.org/ploscompbiol/article?id=10.1371/journal.pcbi.1008274) for their inference and forecasting work.  This model can be expressed in McMasterPandemic with the following.

```{r fig.height=8, fig.width=8}
foi =  (
    vec(c("1", "f")) 
  * struc("beta") 
  * struc("1/N") 
  * (struc("(I) + (E2)") + (struc("f") * struc("(Id) + (E2d)")))
)
params = c(
  N = 5100000,
  D = 5,
  k1 = 0.2,
  k2 = 1,
  q = 0.05,
  ud = 0.1,
  ur = 0.02,
  psir = 0.3,
  shape = 1.73,
  scale = 9.85,
  beta = 0.433,
  f = 1
)
state = c(
  S = 849999,
  E1 = 0.53,
  E2 = 0.13,
  I = 0.67,
  Q = 0,
  R = 0,
  Sd = 4249993,
  E1d = 2.67,
  E2d = 0.67,
  Id = 3.33,
  Qd = 0,
  Rd = 0
)
base_states = names(state)[1:6]
dist_states = base_states %+% "d"
strats = c("", "d") # distancing strategies
ramp_period = seq(from = ymd(20210315), to = ymd(20210322), by = 1)
time_ratio = rev((seq_along(ramp_period) - 1) / (length(ramp_period) - 1))
f2 = 0.22
params_timevar = data.frame(
  Date = ramp_period,
  Symbol = rep("f", length(ramp_period)),
  Value = f2 + time_ratio * (1 - f2),
  Type = rep("abs", length(ramp_period))
)
model = (
  init_model(
    params = params,
    state = state,
    start_date = "2021-02-01",
    end_date = "2021-06-01",
    do_make_state = FALSE,
    do_hazard = TRUE,
    params_timevar = params_timevar
  )
  
  # flow between distancing strategies
  %>% rep_rate(base_states, dist_states, ~ (ud))
  %>% rep_rate(dist_states, base_states, ~ (ur))
  
  # force of infection
  %>% vec_rate(
    "S"  %+% strats, 
    "E1" %+% strats, 
    foi
  )
  
  # flow within distancing strategies
  %>% rep_rate(
    "E1" %+% strats,
    "E2" %+% strats,
    ~ (k1)
  )
  %>% rep_rate(
    "E2" %+% strats,
    "I"  %+% strats,
    ~ (k2)
  )
  %>% rep_rate(
    "I" %+% strats,
    "Q" %+% strats,
    ~ (q)
  )
  %>% rep_rate(
    expand_names(c("I", "Q"), strats, ""),
    expand_names(c("R", "R"), strats, ""),
    ~ (1/D)
  )
  
  # add outflow for all inflows
  %>% add_outflow()
  
  # compute indices to be passed to the tmb engine
  %>% add_tmb_indices()
)
(model
  %>% simulate_state_vector(add_dates = TRUE)
  %>% pivot_longer(!Date)
  %>% ggplot()
   +  facet_wrap(~ name, scales = 'free', ncol = 2)
   +  geom_line(aes(Date, value))
)
```

## Outflows and Accumulators

Up until now we have been assuming that all inflows are associated with a corresponding outflow. The number of individuals that go into a box must exactly balance numbers of individuals that come out of other boxes into this one. There are situations however where one wants to keep track of the number of individuals that have ever entered a particular box. To do this we set up accumulator boxes that have no corresponding outflow, and whose inflows are identical to those of other boxes.

TODO: unfinished

```{r echo = FALSE, eval = FALSE}
plot_sir = function(model) {
  (model
    %>% simulate_state_vector(add_dates = TRUE)
    %>% pivot_longer(!Date)
    %>% rename(state = value, epi_cat = name)
    %>% ggplot
     +  geom_line(aes(x = Date, y = state, colour = epi_cat))
  )
}
```

```{r echo = FALSE, eval = FALSE}
(
  init_model(
    params = c(
      gamma = 1e-1, # per-infected recovery rate
      beta = 5e-9   # per-contact transmission rate
    ),
    state = c(S = 20000, I = 100, R = 0),
    start_date = "2000-01-01",
    end_date = "2000-02-01",
    do_hazard = TRUE,
    do_make_state = FALSE
  )
  %>% add_rate("S", "I", ~ (beta) * (S) * (I))
  %>% add_rate("I", "R", ~ (gamma))
  %>% add_tmb_indices
  %>% plot_sir
)
```

```{r echo = FALSE, eval = FALSE}
(
  init_model(
    params = c(
      gamma = 1e-1, # per-infected recovery rate
      beta = 5e-9   # per-contact transmission rate
    ),
    state = c(S = 20000, I = 100, R = 0, A = 100),
    start_date = "2000-01-01",
    end_date = "2000-02-01",
    do_hazard = FALSE,
    do_make_state = FALSE
  )
  %>% add_rate("S", "I", ~ (beta) * (S) * (I))
  %>% add_rate("S", "A", ~ (beta) * (S) * (I))
  %>% add_rate("I", "R", ~ (gamma))
  %>% add_outflow("S", "I")
  %>% add_outflow("I", "R")
  %>% add_tmb_indices
  %>% plot_sir
)
```

```{r echo = FALSE, eval = FALSE}
sir_model = (
  init_model(
    params = c(
      gamma = 1e-1, # per-infected recovery rate
      beta = 5e-9   # per-contact transmission rate
    ),
    state = c(S = 20000, I = 100, R = 0),
    start_date = "2000-01-01",
    end_date = "2000-02-01",
    do_hazard = TRUE,
    do_make_state = FALSE
  )
  %>% add_rate("S", "I", ~ (beta) * (S) * (I))
  %>% add_rate("I", "R", ~ (gamma))
  %>% add_outflow()
  %>% add_tmb_indices
)
sir_model
```

## Canned Models

### Base MacPan COVID Model -- `make_base_model`

### Vaccination and Variant COVID Model -- `make_vaccination_model`

## Hooking up to Standard MacPan Tools

### Hooking up to `run_sim`

### Hooking up to `calibrate`

### Hooking up to `forecast`
