---
title: "Flexible and Fast Modelling with McMasterPandemic -- Examples"
output: html_document
---

[![Live
Version](https://img.shields.io/static/v1.svg?label=Updated&message=`r Sys.Date()`&color=blue)](https://canmod.net/misc/flex_intro)

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, message=FALSE}
library(ggplot2)
library(McMasterPandemic)
library(dplyr)
library(tidyr)
library(lubridate)
library(equatiomatic)
```

## Hello World: Simulating an SIR Model

```{r}
state = c(S = 20000, I = 100, R = 0)
sir_model = (
  init_model(
    params = c(
      gamma = 0.06,
      beta = 0.15,
      N = sum(state)
    ),
    state = state,
    start_date = "2000-01-01",
    end_date = "2000-05-01",
    do_hazard = TRUE,
    do_make_state = FALSE
  )
  %>% add_rate("S", "I", ~ (1/N) * (beta) * (I))
  %>% add_rate("I", "R", ~ (gamma))
  %>% add_outflow()
  %>% update_tmb_indices
)
sir_model
```

```{r}
(sir_model
    %>% simulate_state_vector(add_dates = TRUE)
    %>% pivot_longer(!Date)
    %>% rename(state = value, epi_cat = name)
    %>% ggplot
     +  geom_line(aes(x = Date, y = state, colour = epi_cat))
  )
```

```{r}
(sir_model
    %>% simulate_changing_ratemat_elements(add_dates = TRUE)
    %>% rename(`force of infection` = S_to_I)
    %>% ggplot
     +  geom_line(aes(x = Date, y = `force of infection`))
  )
```

## Structure: Two-Strain SIR

```{r}

strains = c("wild", "variant")
state = c(
  S = 20000, 
  I_wild = 49, I_variant = 1,
  R_wild = 0,   R_variant = 0
)
two_strain_model = (
  init_model(
    params = c(
      gamma = 0.06,       
      beta_wild = 0.15,   
      beta_variant = 0.25,
      N = sum(state)
    ),
    state = state,
    start_date = "2000-01-01",
    end_date = "2000-05-01",
    do_hazard = TRUE,
    do_make_state = FALSE
  )
  %>% vec_rate(
    "S",
    "I" %_% strains, 
     vec("beta" %_% strains) * struc("1/N") * vec("I" %_% strains)
  )
  %>% rep_rate("I", "R", ~ (gamma))
  %>% add_outflow()
  %>% update_tmb_indices
)
```

```{r}
regex = "^(S|I|R)(_*)(.*)"
(two_strain_model
    %>% simulate_state_vector(add_dates = TRUE)
    %>% pivot_longer(!Date)
    %>% rename(state = value)
    %>% mutate(strain  = sub(pattern = regex, replacement = "\\3", name))
    %>% mutate(epi_cat = sub(pattern = regex, replacement = "\\1", name))
    %>% ggplot
     +  geom_line(aes(x = Date, y = state, colour = epi_cat, linetype = strain))
  )
```

```{r}
(two_strain_model
    %>% simulate_changing_ratemat_elements(add_dates = TRUE)
    %>% pivot_longer(starts_with("S_to_I"))
    %>% mutate(name = sub("S_to_I_", "", name))
    %>% rename(`force of infection` = value)
    %>% rename(strain = name)
    %>% ggplot
     +  geom_line(aes(x = Date, y = `force of infection`, colour = strain))
  )
```

## Erlang SEIR

David, Jonathan, and David describe the [Erlang SEIR model](https://epubs.siam.org/doi/pdf/10.1137/18M1186411) in continuous time. Here is a discrete time version of it.
```{r}
n = 4 # number of I states
m = 6 # number of E states
erlang_seir = (init_model(
  # FIXME: only working for no demography (so mu = 0 for now)
  params = c(mu = 0, beta = 1.5, m = m, n = n,
             gamma = 1.2, sigma = 0.1),
  state = c(
    S = 1-1e-3,
    setNames(rep(0, m), "E" %_% 1:m),
    I_1 = 1e-3,
    setNames(rep(0, n-1), "I" %_% 2:n),
    R = 0,
    D = 0,
    . = 1),
  start_date = "2000-01-01",
  end_date = "2000-04-01",
  do_make_state = FALSE,
  do_hazard = TRUE
)
  %>% add_state_param_sum("I", "^I_[0-9]+")

  # birth
  %>% add_rate(".", "S", ~ (mu))

  # death
  %>% rep_rate(
    from = c("S", "E" %_% 1:m, "I" %_% 1:n, "R"), 
    to = "D",
    formula = ~ (mu))
  
  # infection
  %>% add_rate("S", "E_1", ~ (beta) * (I))

  # become infectious
  %>% add_rate("E" %_% m, "I_1", ~ (m) * (sigma))

  # sojourn through exposed compartments
  %>% rep_rate(
    "E" %_% 1:(m-1),
    "E" %_% 2:m,
    ~ (m) * (sigma)
  )

  # sojourn through infectious compartments
  %>% rep_rate(
    "I" %_% 1:(n-1),
    "I" %_% 2:n,
    ~ (n) * (gamma)
  )

  # recovery
  %>% add_rate("I" %_% n, "R", ~ (n) * (gamma))

  # nothing flows out of . because it is a dummy
  # state used to generate flows that are not per-capita
  %>% add_outflow("[^.]")
  %>% update_tmb_indices
)

(erlang_seir
  %>% simulate_state_vector(add_dates = TRUE)
  %>% select(-., -D)
  %>% pivot_longer(-Date, names_to = "State", values_to = "Count")
  %>% mutate(State = sub("_[0-9]+", "", State))
  %>% group_by(Date, State)
  %>% summarise(Count = sum(Count))
  %>% ungroup
  %>% ggplot() + geom_line(aes(Date, Count, colour = State))
)

# check to make sure that the population density
# remains constant at one
(erlang_seir
  %>% simulate_state_vector(add_dates = FALSE)
  %>% select(-., -D)
  %>% rowSums
  %>% sapply(all.equal, 1L)
  %>% sapply(isTRUE)
  %>% all
)
```

## The SIRV model

```{r}
state = c(S = 20000, I = 100, R = 0, V = 0)
params = c(
      gamma = 0.06,
      beta = 0.15,
      v = 0, # initial vaccination rate
      N = sum(state)
    )

# roll out the vaccine by bumping the 
# vaccination rate twice
params_timevar = data.frame(
  Date = c("2020-02-01", "2020-03-01"), 
  Symbol = c("v", "v"),  
  Value = c(0.01, 0.1), 
  Type = c("abs", "abs"))

sirv_model = (
  init_model(
    params = params,
    state = state,
    params_timevar = params_timevar,
    start_date = "2020-01-01",
    end_date = "2020-05-01",
    do_hazard = TRUE,
    do_make_state = FALSE
  )
  %>% add_rate("S", "I", ~ (beta) * (1/N) * (I))
  %>% add_rate("I", "R", ~ (gamma))
  %>% add_rate("S", "V", ~  (v))
  %>% add_outflow()
  %>% update_tmb_indices
)

(sirv_model
    %>% simulate_state_vector(add_dates = TRUE)
    %>% pivot_longer(!Date)
    %>% rename(state = value, epi_cat = name)
    %>% ggplot
     +  geom_line(aes(x = Date, y = state, colour = epi_cat))
     +  geom_vline(aes(xintercept = as.Date(Date)), data = params_timevar, colour = 'lightgrey')
  )

```


## Variolation model

```{r}
state = c(
  S = 20000, 
  I_severe = 50, I_mild = 50,
  R_mild = 0,   R_severe = 0
)
params = c(
  nu = 0.0105, # birth rate
  mu = 0.0105, # death rate
  delta = 1/7, # waning immunity rate
  gamma_mild = 1/(3.3 + 9),  # recovery rate of mild cases
  gamma_severe = 1/(3.3 + 14), # recovery rate of severe cases
  beta_mild = 0.15,  # transmission rate of infections leading to mild cases
  beta_severe = 0.3, # transmission rate of infections leading to severe cases
  m = 0.6 # probability of developing mild illness
)

# model structure
all_states = names(state)
severity = c("mild", "severe")
beta_vec = vec("beta" %_% severity)
I_vec = vec("I" %_% severity)
foi = kronecker(vec("m", "1-m"), sum(beta_vec * I_vec * struc("1/N")))

variolation_model <- (
  init_model(
    params = params,
    state = state,
    start_date = "2020-01-01",
    end_date = "2020-04-01",
    do_hazard = FALSE,
    do_make_state = FALSE
  )
  %>% add_state_param_sum("N", any_var(state))
  
  # births and deaths 
  # (FIXME: this only works if nu = mu)
  %>% rep_rate(
    all_states,
    "S",
    ~ (nu)
  )
  
  # infection
  %>% vec_rate(
    "S",
    "I" %_% severity, 
     foi
  )
  
  # waning immunity
  %>% rep_rate(
    "R" %_% severity,
    "S",
    ~ (delta)
  )
  
  # recovery
  %>% vec_rate(
    "I" %_% severity,
    "R" %_% severity,
    vec("gamma" %_% severity)
  )
  
  %>% add_outflow()
  %>% update_tmb_indices
)

(variolation_model
    %>% simulate_state_vector(add_dates = TRUE)
    %>% pivot_longer(!Date)
    %>% rename(state = value, epi_cat = name)
    %>% ggplot
     +  geom_line(aes(x = Date, y = state, colour = epi_cat))
  )

```

## SEIRD Model

This is the [Mac Theo Bio Model](https://github.com/mac-theobio/mac-theobio.github.io/blob/master/seird.jpg).

```{r}
state = c(S = 20000, E = 50, I = 50, R = 0, D = 0, Null = 0)
params = c(
  # transmission rates of susceptibles from live and dead individuals
  beta_I = 5,
  beta_D = 2,
  
  # average times in various boxes
  T_E = 10,
  T_I = 14,
  T_D = 10,
  
  # probability of death given infection
  f = 0.5
)
seird_model = (
  init_model(
    params = params, 
    state = state,
    start_date = "2000-01-01",
    end_date = "2000-03-01",
    do_hazard = TRUE,
    do_make_state = FALSE
  )
  %>% add_state_param_sum("N", any_var(state))
  %>% add_rate("S", "E", ~ (beta_I) * (1/N) * (I) + (beta_D) * (1/N) * (D))
  %>% add_rate("E", "I", ~ (1/T_E))
  %>% add_rate("I", "R", ~ (1/T_I) * (1 - f))
  %>% add_rate("I", "D", ~ (1/T_I) * (f))
  %>% add_rate("D", "Null", ~ (1/T_D))
  %>% add_outflow()
  %>% update_tmb_indices
)
(seird_model
    %>% simulate_state_vector(add_dates = TRUE)
    %>% pivot_longer(!Date)
    %>% rename(state = value, epi_cat = name)
    %>% ggplot
     +  geom_line(aes(x = Date, y = state, colour = epi_cat))
  )
```

## Covid SEIR

The BC covid modelling group uses this [compartmental model](https://journals.plos.org/ploscompbiol/article?id=10.1371/journal.pcbi.1008274) for their inference and forecasting work.  This model can be expressed in McMasterPandemic with the following.

*DISCLAIMER: This is not meant to illustrate a realistic forecast, but rather to illustrate how the basic model structure can be expressed in McMasterPandemic*
```{r fig.height=8, fig.width=8}
foi =  (
    vec(c("1", "f")) 
  * struc("beta") 
  * struc("1/N") 
  * (struc("(I) + (E2)") + (struc("f") * struc("(Id) + (E2d)")))
)
params = c(
  N = 5100000,
  D = 5,
  k1 = 0.2,
  k2 = 1,
  q = 0.05,
  ud = 0.1,
  ur = 0.02,
  psir = 0.3,
  shape = 1.73,
  scale = 9.85,
  beta = 0.433,
  f = 1
)
state = c(
  S = 849999,
  E1 = 0.53,
  E2 = 0.13,
  I = 0.67,
  Q = 0,
  R = 0,
  Sd = 4249993,
  E1d = 2.67,
  E2d = 0.67,
  Id = 3.33,
  Qd = 0,
  Rd = 0
)
base_states = names(state)[1:6]
dist_states = base_states %+% "d"
strats = c("", "d") # distancing strategies
ramp_period = seq(from = ymd(20210315), to = ymd(20210322), by = 1)
time_ratio = rev((seq_along(ramp_period) - 1) / (length(ramp_period) - 1))
f2 = 0.22
params_timevar = data.frame(
  Date = ramp_period,
  Symbol = rep("f", length(ramp_period)),
  Value = f2 + time_ratio * (1 - f2),
  Type = rep("abs", length(ramp_period))
)
model = (
  init_model(
    params = params,
    state = state,
    start_date = "2021-02-01",
    end_date = "2021-06-01",
    do_make_state = FALSE,
    do_hazard = TRUE,
    params_timevar = params_timevar
  )
  
  # flow between distancing strategies
  %>% rep_rate(base_states, dist_states, ~ (ud))
  %>% rep_rate(dist_states, base_states, ~ (ur))
  
  # force of infection
  %>% vec_rate(
    "S"  %+% strats, 
    "E1" %+% strats, 
    foi
  )
  
  # flow within distancing strategies
  %>% rep_rate(
    "E1" %+% strats,
    "E2" %+% strats,
    ~ (k1)
  )
  %>% rep_rate(
    "E2" %+% strats,
    "I"  %+% strats,
    ~ (k2)
  )
  %>% rep_rate(
    "I" %+% strats,
    "Q" %+% strats,
    ~ (q)
  )
  %>% rep_rate(
    expand_names(c("I", "Q"), strats, ""),
    expand_names(c("R", "R"), strats, ""),
    ~ (1/D)
  )
  
  # add outflow for all inflows
  %>% add_outflow()
  
  # compute indices to be passed to the tmb engine
  %>% update_tmb_indices()
)
(model
  %>% simulate_state_vector(add_dates = TRUE)
  %>% pivot_longer(!Date)
  %>% ggplot()
   +  facet_wrap(~ name, scales = 'free', ncol = 2)
   +  geom_line(aes(Date, value))
)
```

## BC Covid Omicron

The BC group also developed a model with two strains for the Omicron wave of Covid-19.

*DISCLAIMER: This is not meant to illustrate a realistic forecast, but rather to illustrate how the basic model structure can be expressed in McMasterPandemic*

```{r make_init_omicron, echo=FALSE}

make_init = function(params,
                     N=5.07e6, vaxlevel = 0.8,
                      port_wane = 0.04,
                      past_infection = 0.1,
                      incres = 300,
                      incmut = 10) {
  params_list = as.list(params)
  ff=6/7 # fudge factor . hard to get incidence right since it depends on other parameters too (2/3)
  Vtot = vaxlevel*N*(1-port_wane) # allocate to V, Ev, Iv
  Wtot = vaxlevel*N*port_wane # allocate to W, Ew, Iw
  # some have had covid. but they might also have been vaccinated.
  # set up Rs
  R0 = N*past_infection*(1-vaxlevel)  # past infections, but not vaccinated
  Rv0 = N*past_infection*(vaxlevel) * (1-port_wane) # recovered, vaxd and not waned
  Rw0 = N*past_infection*(vaxlevel) * port_wane # rec, vaxd, waned
  # set up Es : resident
  Ertot = ff*incres/params_list$sigma # total Er
  Ervw = vaxlevel*params_list$ve*Ertot # vaccinated
  Erw0 = Ervw * port_wane # waned
  Erv0 = Ervw *(1-port_wane) # not waned.  these two add to Ervwboth
  Er0 = (1-vaxlevel*params_list$epsilon_r)*Ertot # unvaccinated
  # set up Es : mutant
  Emtot = ff*incmut/params_list$sigma # total Er
  Emvw = vaxlevel*params_list$ve*Emtot # vaccinated
  Emw0 = Emvw * port_wane # waned
  Emv0 = Emvw *(1-port_wane) # not waned.  these two add to Ervwboth
  Em0 = (1-vaxlevel*params_list$epsilon_m)*Emtot # unvaccinated
  # set up Is. for now just make them 2x Es since they last twice as long
  Ir0=ff*2*Er0; Irv0 = ff*2*Erv0; Irw0 = ff*2*Erw0
  Im0 = ff*2*Em0; Imv0=ff*2*Emv0; Imw0= ff*2*Emw0

  # set up V0, W0
  # first line plus Rv0 adds to the V total but we have to leave room for the E, I
  V0 = N*(1-past_infection)*vaxlevel*(1-port_wane) -
    (Erv0+Irv0+Emv0+Imv0)*(1-port_wane)
  W0 = N*(1-past_infection)*vaxlevel*(port_wane) -
    ( Erw0+Irw0+Emw0+Imw0)*(port_wane)
  # set up S
  initmost = c(Er=Er0, Em =Em0, Ir=Ir0, Im=Im0, R=R0,
               V=V0, Erv=Erv0, Emv=Emv0, Irv = Irv0,  Imv=Imv0, Rv=Rv0,
               W=W0, Erw=Erw0, Emw=Emw0, Irw = Irw0,  Imw=Imw0, Rw=Rw0)
  state = c(S=N-sum(initmost), initmost)
  return(state)
}
```
```{r omicron, fig.height=8, fig.width=8}
params = c(
  sigma=1/3, # incubation period (3 days) (to fixed)
  gamma=1/(5), #recovery rate (fixed)
  nu =0.007, #vax rate: 0.7% per day (fixed)
  mu=1/(82*365), # 1/life expectancy (fixed)
  w1 = 1/(3*365),# waning rate from R to S (fixed)
  w2 = 1/(3*365), # waning rate from Rv to V (fixed)
  w3 = 1/(3*365),# waning rate Rw to W (fixed)
  ve=1, # I think this should be 1. it is not really efficacy  ( fixed)
  beta_r=0.72, #transmission rate (to estimate) (0.35)
  beta_m=0.8*2.2, #transmission rate (to estimate)(*1.9)
  epsilon_r = (1-0.8), # % this should be 1-ve
  epsilon_m = (1-0.6), # % escape capacity #(fixed)
  b= 0.006, # booster rate  (fixed)
  beff = 0.7, # booster efficacy
  wf=0.2, # protection for newly recoverd #0.2
  N=5e6,
  E0=5,
  S0=1-1e-5,
  c=1
)

# dimensions of model structure
vax_states = c("unvax", "onedose", "boost")
variant_states = c("delta", "omicron")
epi_states = c("S", "E" %_% variant_states, "I" %_% variant_states, "R")

# vectors representing variant model structure
I_delta   = vec("I" %_% "delta"   %_% vax_states)
I_omicron = vec("I" %_% "omicron" %_% vax_states)
E_delta   = vec("E" %_% "delta"   %_% vax_states)
E_omicron = vec("E" %_% "omicron" %_% vax_states)

# initial state vector
state = layered_zero_state(epi_states, vax_states)
all_states = names(state)
state[] = unname(make_init(params))
print(state)

two_strain_bc = (init_model(
    params = params,
    state = state,
    start_date = "2022-01-01",
    end_date = "2022-05-01",
    do_make_state = FALSE,
    do_hazard = TRUE
  )

  # sum over every vax status to get
  # total numbers in I boxes for each
  # variant
  %>% add_state_param_sum("I_delta", "I_delta_")
  %>% add_state_param_sum("I_omicron", "I_omicron_")

  # R to S boxes for every vax status
  #   -- waning
  %>% vec_rate(
    from = "R" %_% vax_states,
    to   = "S" %_% vax_states,
    vec('w1', 'w2', 'w3')
  )

  # S_delta to E_delta for every vax status
  #   -- delta force of infection
  %>% vec_rate(
    from = "S"             %_% vax_states,
    to   = "E" %_% "delta" %_% vax_states,
    struc("(c) * (beta_r) * (1/N) * (I_delta)") * vec("1", "epsilon_r", "epsilon_r")
  )

  # S_omicron to E_omicron for every vax status
  #   -- omicron force of infection
  %>% vec_rate(
    from = "S"               %_% vax_states,
    to   = "E" %_% "omicron" %_% vax_states,
    struc("(c) * (beta_m) * (1/N) * (I_omicron)") * vec("1", "epsilon_m", "epsilon_m")
  )

  # R to E_delta for every vax status
  #   -- delta force of infection of recovered individuals
  %>% rep_rate(
    from = "R"             %_% vax_states,
    to   = "E" %_% "delta" %_% vax_states,
    ~ (wf) * (epsilon_r) * (c) * (beta_r) * (1/N) * (I_delta)
  )

  # R to E_omicron for every vax status
  #   -- omicron force of infection of recovered individuals
  %>% rep_rate(
    from = "R"               %_% vax_states,
    to   = "E" %_% "omicron" %_% vax_states,
    ~ (wf) * (epsilon_m) * (c) * (beta_m) * (1/N) * (I_omicron)
  )

  # E to I for all variant-vax combinations
  %>% rep_rate(
    from = "E" %_% expand_names(variant_states, vax_states),
    to   = "I" %_% expand_names(variant_states, vax_states),
    ~ (sigma)
  )

  # recovery for all variant-vax combinations
  %>% rep_rate(
    from = "I" %_% expand_names(variant_states, vax_states),
    to   = "R" %_% rep(vax_states, each = length(variant_states)),
    ~ (gamma)
  )

  # demographics
  %>% rep_rate(
    from = all_states,
    to   = "S_unvax",
    ~ (mu)
  )

  # vaccination
  %>% add_rate(from = "S_unvax",   to = "S_onedose", ~ (nu) * (ve))
  %>% add_rate(from = "S_onedose", to = "S_boost",   ~ (b)  * (ve))

  %>% add_outflow
  %>% update_tmb_indices
)

category_pattern = "^(S|E|I|R)(_omicron|_delta)?_(unvax|onedose|boost)$"
(two_strain_bc
  %>% simulate_state_vector(add_date = TRUE, format = 'long')
  %>% mutate(`Epi Status` = sub(category_pattern, '\\1\\2', Compartment, perl = TRUE))
  %>% mutate(`Vaccination Status` = sub(category_pattern, '\\3', Compartment, perl = TRUE))
  %>% arrange(`Epi Status`, `Vaccination Status`)
  %>% mutate(Compartment = factor(Compartment, all_states))
  %>% ggplot()
   +  facet_wrap(
     ~ Compartment, ncol = 3, scales = 'free', dir = 'v')
   +  geom_line(aes(Date, State))
)
```

## Outflows and Accumulators

Up until now we have been assuming that all inflows are associated with a corresponding outflow. The number of individuals that go into a box must exactly balance numbers of individuals that come out of other boxes into this one. There are situations however where one wants to keep track of the number of individuals that have ever entered a particular box. To do this we set up accumulator boxes that have no corresponding outflow, and whose inflows are identical to those of other boxes.

TODO: unfinished

```{r echo = FALSE, eval = FALSE}
plot_sir = function(model) {
  (model
    %>% simulate_state_vector(add_dates = TRUE)
    %>% pivot_longer(!Date)
    %>% rename(state = value, epi_cat = name)
    %>% ggplot
     +  geom_line(aes(x = Date, y = state, colour = epi_cat))
  )
}
```

```{r echo = FALSE, eval = FALSE}
(
  init_model(
    params = c(
      gamma = 1e-1, # per-infected recovery rate
      beta = 5e-9   # per-contact transmission rate
    ),
    state = c(S = 20000, I = 100, R = 0),
    start_date = "2000-01-01",
    end_date = "2000-02-01",
    do_hazard = TRUE,
    do_make_state = FALSE
  )
  %>% add_rate("S", "I", ~ (beta) * (S) * (I))
  %>% add_rate("I", "R", ~ (gamma))
  %>% update_tmb_indices
  %>% plot_sir
)
```

```{r echo = FALSE, eval = FALSE}
(
  init_model(
    params = c(
      gamma = 1e-1, # per-infected recovery rate
      beta = 5e-9   # per-contact transmission rate
    ),
    state = c(S = 20000, I = 100, R = 0, A = 100),
    start_date = "2000-01-01",
    end_date = "2000-02-01",
    do_hazard = FALSE,
    do_make_state = FALSE
  )
  %>% add_rate("S", "I", ~ (beta) * (S) * (I))
  %>% add_rate("S", "A", ~ (beta) * (S) * (I))
  %>% add_rate("I", "R", ~ (gamma))
  %>% add_outflow("S", "I")
  %>% add_outflow("I", "R")
  %>% update_tmb_indices
  %>% plot_sir
)
```

```{r echo = FALSE, eval = FALSE}
sir_model = (
  init_model(
    params = c(
      gamma = 1e-1, # per-infected recovery rate
      beta = 5e-9   # per-contact transmission rate
    ),
    state = c(S = 20000, I = 100, R = 0),
    start_date = "2000-01-01",
    end_date = "2000-02-01",
    do_hazard = TRUE,
    do_make_state = FALSE
  )
  %>% add_rate("S", "I", ~ (beta) * (S) * (I))
  %>% add_rate("I", "R", ~ (gamma))
  %>% add_outflow()
  %>% update_tmb_indices
)
sir_model
```

## LaTeX Engine

```{r, results='asis', eval=FALSE}
library(McMasterPandemic)
library(dplyr)
params = read_params("PHAC.csv")
state = make_state(params = params)
model = make_base_model(
  params = params, state = state,
  start_date = "2000-01-01", end_date = "2000-02-01"
)
latex = make_latex_flows(model)
cat(format("$" %+% latex %+% "$"), sep = "\n\n")
```

## Canned Models

### Base MacPan COVID Model -- `make_base_model`

### Vaccination and Variant COVID Model -- `make_vaccination_model`

## Hooking up to Standard MacPan Tools

### Hooking up to `run_sim`

### Hooking up to `calibrate`

### Hooking up to `forecast`
