---
title: "Flexible and Fast Modelling with McMasterPandemic"
output: html_document
---

[![Live
Version](https://img.shields.io/static/v1.svg?label=Updated&message=`r Sys.Date()`&color=blue)](https://canmod.net/misc/flex_intro)

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, message=FALSE}
library(McMasterPandemic)
library(dplyr)
library(tidyr)
library(ggplot2)
```

## Hello World: Simulating an SIR Model

```{r}
sir_model = (
  init_model(
    params = c(
      gamma = 1e-1, # per-infected recovery rate
      beta = 5e-9   # per-contact transmission rate
    ),
    state = c(S = 20000, I = 100, R = 0),
    start_date = "2000-01-01",
    end_date = "2000-02-01",
    do_hazard = TRUE,
    do_make_state = FALSE
  )
  %>% add_rate("S", "I", ~ (beta) * (S) * (I))
  %>% add_rate("I", "R", ~ (gamma))
  %>% add_outflow()
  %>% add_tmb_indices
)
sir_model
```

```{r}
(sir_model
    %>% simulate_state_vector(add_dates = TRUE)
    %>% pivot_longer(!Date)
    %>% rename(state = value, epi_cat = name)
    %>% ggplot
     +  geom_line(aes(x = Date, y = state, colour = epi_cat))
  )
```

```{r}
(sir_model
    %>% simulate_changing_ratemat_elements(add_dates = TRUE)
    %>% rename(`force of infection` = S_to_I)
    %>% ggplot
     +  geom_line(aes(x = Date, y = `force of infection`))
  )
```

## Structure: Two-Strain SIR

```{r}

strains = c("wild", "variant")

two_strain_model = (
  init_model(
    params = c(
      gamma = 1e-1,       # per-infected recovery rate
      beta_wild = 5e-9,   # per-contact transmission rate for wild type
      beta_variant = 6e-9 # per-contact transmission rate for variant
    ),
    state = c(
      S = 20000, 
      I_wild = 50, I_variant = 50,
      R_wild = 0,   R_variant = 0
    ),
    start_date = "2000-01-01",
    end_date = "2000-02-01",
    do_hazard = TRUE,
    do_make_state = FALSE
  )
  %>% vec_rate(
    rep("S", length(strains)),
    "I" %_% strains, 
     vec("beta" %_% strains) * struc("S") * vec("I" %_% strains)
  )
  %>% rep_rate("I", "R", ~ (gamma))
  %>% add_outflow()
  %>% add_tmb_indices
)
```

```{r}
regex = "^(S|I|R)(_*)(.*)"
(two_strain_model
    %>% simulate_state_vector(add_dates = TRUE)
    %>% pivot_longer(!Date)
    %>% rename(state = value)
    %>% mutate(strain  = sub(pattern = regex, replacement = "\\3", name))
    %>% mutate(epi_cat = sub(pattern = regex, replacement = "\\1", name))
    %>% ggplot
     +  geom_line(aes(x = Date, y = state, colour = epi_cat, linetype = strain))
  )
```

```{r}
(two_strain_model
    %>% simulate_changing_ratemat_elements(add_dates = TRUE)
    %>% pivot_longer(starts_with("S_to_I"))
    %>% mutate(name = sub("S_to_I_", "", name))
    %>% rename(`force of infection` = value)
    %>% rename(strain = name)
    %>% ggplot
     +  geom_line(aes(x = Date, y = `force of infection`, colour = strain))
  )
```

## The SIRV model

```{r}

sirv_model = (
  init_model(
    params = c(
      gamma = 1e-2, # per-infected recovery rate
      beta = 2e-9,  # per-contact transmission rate
      v = 1e-5),
    params_timevar = data.frame(Date = c("2020-01-05", "2020-01-10"), Symbol = c("v", "v"),  Value = c(0.1, 2), Type = c("rel_orig", "rel_orig")),
    state = c(S = 20000, I = 100, R = 0, V = 0),
    start_date = "2020-01-01",
    end_date = "2020-05-01",
    do_hazard = FALSE,
    do_make_state = FALSE
  )
  %>% add_rate("S", "I", ~ (beta) * (S) * (I))
  %>% add_rate("I", "R", ~ (gamma))
  %>% add_rate("S", "V", ~  (v) * (S))
  %>% add_outflow("S|I|R|V", "S|I|R|V")
  %>% add_tmb_indices
)

(sirv_model
    %>% simulate_state_vector(add_dates = TRUE)
    %>% pivot_longer(!Date)
    %>% rename(state = value, epi_cat = name)
    %>% ggplot
     +  geom_line(aes(x = Date, y = state, colour = epi_cat))
  )

```


## Variolation model

```{r}
kinds = c("mild", "severe")
variolation_model <- (
  init_model(
    params = c(
      rho = 0.1, #probability of mild infection
      gamma_mild = 1e-1,       # per-infected recovery rate
      beta_mild = 5e-9,   # per-contact transmission rate for wild type
      gamma_severe = 2e-1,
      beta_severe = 6e-9 # per-contact transmission rate for variant
    ),
    params_timevar = data.frame(Date = c("2020-01-05", "2020-01-07"), Symbol = c("rho", "beta_mild"),  Value = c(0.9, 3), Type = c("abs", "rel_orig")),    state = c(
      S = 20000, 
      I_severe = 50, I_mild = 50,
      R_mild = 0,   R_severe = 0
    ),
    start_date = "2020-01-01",
    end_date = "2020-05-01",
    do_hazard = FALSE,
    do_make_state = FALSE
  )
  %>% add_rate("S", "I_mild", ~ (beta_mild) * (S) * (rho)* (I_mild) + (beta_mild) * (S) * (rho)* (I_severe))
  %>% add_rate("S", "I_severe", ~ (beta_mild) * (S) * (1-rho)* (I_mild) + (beta_mild) * (S) * (1-rho)* (I_severe))
  %>% add_rate("I_mild", "R_mild", ~ (gamma_severe))
  %>% add_rate("I_severe", "R_severe", ~ (gamma_mild))
  %>% add_outflow("(S|I_mild|R_mild|I_severe|R_severe)", "(S|I_mild|R_mild|I_severe|R_severe)")
  %>% add_tmb_indices
)

(variolation_model
    %>% simulate_state_vector(add_dates = TRUE)
    %>% pivot_longer(!Date)
    %>% rename(state = value, epi_cat = name)
    %>% ggplot
     +  geom_line(aes(x = Date, y = state, colour = epi_cat))
  )

```

``
## SEIRD Model

This is the [Mac Theo Bio Model](https://github.com/mac-theobio/mac-theobio.github.io/blob/master/seird.jpg).

```{r}
seird_model = (
  init_model(
    params = c(
      beta_I = 1e-5,
      beta_D = 1e-6,
      T_E = 1e-7,
      T_I = 1e-8,
      f = 0.5,
      T_D = 1e-4
    ),
    state = c(S = 20000, E = 50, I = 50, R = 0, D = 0, Gone = 0),
    start_date = "2000-01-01",
    end_date = "2000-02-01",
    do_hazard = TRUE,
    do_make_state = FALSE
  )
  %>% add_rate("S", "E", ~ (beta_I) * (S) * (I) + (beta_D) * (S) * (D))
  %>% add_rate("E", "I", ~ (T_E))
  %>% add_rate("I", "R", ~ (T_I) * (1 - f))
  %>% add_rate("I", "D", ~ (T_I) * (f))
  %>% add_rate("D", "Gone", ~ (T_D))
  %>% add_outflow(".*", ".*")
  %>% add_tmb_indices
)
seird_model
```

## Sums of State Variables and Parameters

```{r}
sir_model = (
  init_model(
    params = c(
      gamma = 1e-1, # per-infected recovery rate
      beta = 5e-9   # per-contact transmission rate
    ),
    state = c(S = 20000, I = 100, R = 0),
    start_date = "2000-01-01",
    end_date = "2000-02-01",
    do_hazard = TRUE,
    do_make_state = FALSE
  )
  %>% add_rate("S", "I", ~ (beta) * (S) * (I))
  %>% add_rate("I", "R", ~ (gamma))
  %>% add_outflow()
  %>% add_tmb_indices
)
sir_model
```

## Outflows

```{r echo = FALSE, eval = FALSE}
plot_sir = function(model) {
  (model
    %>% simulate_state_vector(add_dates = TRUE)
    %>% pivot_longer(!Date)
    %>% rename(state = value, epi_cat = name)
    %>% ggplot
     +  geom_line(aes(x = Date, y = state, colour = epi_cat))
  )
}
```

```{r echo = FALSE, eval = FALSE}
(
  init_model(
    params = c(
      gamma = 1e-1, # per-infected recovery rate
      beta = 5e-9   # per-contact transmission rate
    ),
    state = c(S = 20000, I = 100, R = 0),
    start_date = "2000-01-01",
    end_date = "2000-02-01",
    do_hazard = TRUE,
    do_make_state = FALSE
  )
  %>% add_rate("S", "I", ~ (beta) * (S) * (I))
  %>% add_rate("I", "R", ~ (gamma))
  %>% add_tmb_indices
  %>% plot_sir
)
```

```{r echo = FALSE, eval = FALSE}
(
  init_model(
    params = c(
      gamma = 1e-1, # per-infected recovery rate
      beta = 5e-9   # per-contact transmission rate
    ),
    state = c(S = 20000, I = 100, R = 0, A = 100),
    start_date = "2000-01-01",
    end_date = "2000-02-01",
    do_hazard = FALSE,
    do_make_state = FALSE
  )
  %>% add_rate("S", "I", ~ (beta) * (S) * (I))
  %>% add_rate("S", "A", ~ (beta) * (S) * (I))
  %>% add_rate("I", "R", ~ (gamma))
  %>% add_outflow("S", "I")
  %>% add_outflow("I", "R")
  %>% add_tmb_indices
  %>% plot_sir
)
```

```{r echo = FALSE, eval = FALSE}
sir_model = (
  init_model(
    params = c(
      gamma = 1e-1, # per-infected recovery rate
      beta = 5e-9   # per-contact transmission rate
    ),
    state = c(S = 20000, I = 100, R = 0),
    start_date = "2000-01-01",
    end_date = "2000-02-01",
    do_hazard = TRUE,
    do_make_state = FALSE
  )
  %>% add_rate("S", "I", ~ (beta) * (S) * (I))
  %>% add_rate("I", "R", ~ (gamma))
  %>% add_outflow()
  %>% add_tmb_indices
)
sir_model
```

## Canned Models

### Base MacPan COVID Model -- `make_base_model`

### Vaccination and Variant COVID Model -- `make_vaccination_model`

## Hooking up to Standard MacPan Tools

### Hooking up to `run_sim`

### Hooking up to `calibrate`

### Hooking up to `forecast`
