---
title: "Speedup Comparisons"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Theoretical Considerations

Split up the code executed for a particular MacPan job into three sets:

1. The code that has already been refactored in C++ (e.g. sharcnet round 1)
2. The code that we would like to refactor in C++ (e.g. sharcnet round 2)
3. The code that we will not refactor (e.g. plotting, index construction, argument checking)

Let $r_i$ be the non-refactored run-time of the code in set $i$, and $c_i$ be the refactored run-time.

The total run-time before refactoring is,
$$
t_0 = r_1 + r_2 + r_3
$$
after the first sharcnet round is
$$
t_1 = c_1 + r_2 + r_3
$$
and after the second round is
$$
t_2 = c_1 + c_2 + r_3
$$

Three dimensionless measures of this system are:

* Realized speedup: $\tau_i = \frac{t_0}{t_i}$
* Naive speedup: $\alpha_i = \frac{r_i}{c_i}$
* Code fraction: $\rho_i = \frac{r_i}{t_0}$

The realized speedups can be expressed as:

$$
\tau_1 = \frac{1}{\frac{\rho_1}{\alpha_1} + 1 - \rho_1}
$$
and 
$$
\tau_2 = \frac{1}{\frac{\rho_1}{\alpha_1} + \frac{\rho_2}{\alpha_2} + 1 - \rho_1 - \rho_2}
$$
The first of these is [Amdahl's Law](https://en.wikipedia.org/wiki/Amdahl%27s_law).

For many examples after the first round, we have found realized speedups $\tau_1 \approx 6$ and naive speedups $\alpha_1 \approx 60$. Rearranging Amdahl's Law we can use these numbers to estimate the fraction of the program that was rewritten in terms of the dimensionless time scale, $\rho_1$.

$$
\rho_1 = \frac{\frac{1}{\tau_1} - 1}{\frac{1}{\alpha_1} - 1} \approx 0.85
$$

We can now make projections for the speedups we can expect in round 2, under various assumptions. If we assume equal naive speedups in the two rounds, $\alpha_1 = \alpha_2 = \alpha$, then we can plot the realized speedup as a function of the fraction of the code that we refactor.

$$
\tau_2 = \frac{1}{\frac{1 - \rho_3}{\alpha} + \rho_3}
$$

```{r, echo = FALSE}
alpha_1 = 60
tau_1 = 6
rho_1 = ((1/tau_1) - 1) / ((1/alpha_1) - 1)

rho_3 = seq(0, 1 - rho_1, length = 100)
rho_2 = seq(0, 1 - rho_1, length = 100)
tau_2 = 1 / (((1 - rho_3) / alpha_1) + rho_3)
tau_2 = 1 / (((rho_1 + rho_2) / alpha_1) + 1 - rho_1 - rho_2)

plot(rho_2, tau_2, type = 'l', xlab = "Fraction Refactored in Round 2 (rho_2)", ylab = "Speedup after Round 2 (tau_2)", las = 1, ylim = c(0, alpha_1))
abline(h = c(alpha_1, tau_1), lty = 2)
```

## Benchmarks

### Hardware Overview

```
Model Name:	MacBook Pro
Model Identifier:	MacBookPro16,2
Processor Name:	Quad-Core Intel Core i7
Processor Speed:	2.3 GHz
Number of Processors:	1
Total Number of Cores:	4
L2 Cache (per Core):	512 KB
L3 Cache:	8 MB
Hyper-Threading Technology:	Enabled
Memory:	32 GB
```

### Summarized Results

### Detailed Results

#### Simulating the Basic Model

#### Simulating the Basic Model with a Breakpoint at each Step

#### Calibrating the Basic Model

#### Calibrating Time-Varying Parameters in the Basic Model

#### Calibrating Time-Varying Parameters in the Breakpoint Model

#### Simulating the Vax-Var Model

#### Simulating the Vax-Var-Break Model

#### Calibrating the Vax-Var Model

```{r eval = FALSE}
start_date <- "2020-02-01"
end_date <- "2020-09-01"
options(macpan_pfun_method = "grep")
options(MP_use_state_rounding = FALSE)
options(MP_vax_make_state_with_hazard = FALSE)

params <- read_params("ICU1.csv")
state <- make_state(params = params)
vax_params <- expand_params_vax(
  params = params,
  model_type = "twodose"
)
vax_state <- expand_state_vax(
  x = state,
  model_type = "twodose",
  unif = FALSE
)
params_timevar <- data.frame(
  Date = c(as.Date(start_date) + 30,
           as.Date(start_date) + 60),
  Symbol = c("vax_prop_first_dose", "beta0"),
  Value = rep(0.5, 2),
  Type = rep("rel_orig", 2)
)

synth_reports <- (run_sim(
    params = vax_params,
    start_date = start_date,
    end_date = end_date,
    params_timevar = params_timevar
  )
  %>% mutate(value=round(report), var="report")
  %>% select(date, value, var)
  %>% na.omit()
)

opt_pars <- list(params = c(beta0 = 0.6))

test_model <- make_vaccination_model(
  params = expand_params_S0(vax_params, 1-1e-5),
  state = vax_state,
  params_timevar = params_timevar,
  start_date = start_date, end_date = end_date,
  step_args = list(do_hazard = TRUE)
)

time_wrap(
  fitted_mod_tmb <- calibrate(
    base_params = expand_params_S0(vax_params, 1-1e-5),
    data = synth_reports,
    opt_pars = opt_pars,
    debug = TRUE,
    time_args = list(
      params_timevar = params_timevar
    ),
    sim_args = list(
      ndt = 1,
      step_args = list(do_hazard = TRUE),
      use_flex = TRUE,
      flexmodel = test_model,
      obj_fun = tmb_fun(test_model)
    )
  ),
  fitted_mod <- calibrate(
    base_params = vax_params,
    data = synth_reports,
    opt_pars = opt_pars,
    debug = TRUE,
    time_args = list(
      params_timevar = params_timevar
    ),
    sim_args = list(
      ndt = 1,
      step_args = list(do_hazard = TRUE)
    )
  )
)

expect_equal(fitted_mod$mle2@coef, fitted_mod_tmb$mle2@coef)
expect_equal(fitted_mod$mle2@min, fitted_mod_tmb$mle2@min)
expect_equal(fitted_mod$mle2@vcov, fitted_mod_tmb$mle2@vcov)
```

```
TMB: Time difference of 7.78079 secs
R: Time difference of 1741.036 secs
Realized Speedup: 223.7608

Coefficients:
      params.beta0 log_nb_disp.report 
         0.9994531         18.0142768 
Log-likelihood: -264.88 

expect_equal(fitted_mod$mle2@coef, fitted_mod_tmb$mle2@coef) # PASS
expect_equal(fitted_mod$mle2@min, fitted_mod_tmb$mle2@min)   # PASS
expect_equal(fitted_mod$mle2@vcov, fitted_mod_tmb$mle2@vcov) # PASS
```

#### Calibrating the Vax-Var-Break Model

#### Calibrating the Ontario Model

#### Forecasting the Ontario Model
