---
title: "age examples"
---

**tl;dr** 

- *Good news*: we can build structures that seem to make sense, and basic stuff runs without crashing
- *Bad news*: need to spend more time thinking about scaling/normalization of contact matrices, age structure etc.. Lots more testing to do.

```{r pkgs,message=FALSE}
library(McMasterPandemic)
library(Matrix)
library(cowplot)
library(tidyverse)
library(directlabels)
```

Read regular parameters, make normal (non-testify, non-expanded) parameters

```{r utils, echo=FALSE}
tot_I <- function(x) sum(x[grep("^I[a-z]",names(x))])
## nice plot of sparse matrices
ifun <- function(M,sub="") {
    Matrix::image(Matrix(M),scales=list(y=list(at=seq(nrow(M)),labels=rownames(M)),
                                        x=list(at=seq(ncol(M)),labels=colnames(M), rot=90)),
                  xlab="",ylab="",
                  sub=sub)
}
## drop reporting-related columns (since they're missing in ageified sims)
droprep <- . %>% select(-c(foi,incidence,report,cumRep))
## combine sims and compute differences
mdiff_fun  <- function(L) {
    (map_dfr(L,
             ~ droprep(.) %>% pivot(), ## piping to droprep doesn't work ...
             .id="model")
        %>% group_by(date,var)
        %>% summarise(diff=abs(value-value[1])[-1],
                      lab=sprintf("abs(%s-%s)",model,model[1])[-1],
                      .groups="drop")
    )
}
```

```{r setup1}
## set testing to zero so we don't get testing included to start
pp <- update(read_params("PHAC_testify.csv"), testing_intensity=0)
ss <- make_state(params=pp)
```

Expand the state vector by age categories (by default, ten-year bins up to 91+)
```{r expand_state}
ss2 <- expand_stateval_age(ss)
aa <- mk_agecats() ## default
## hack so we have an infective
ss2[grep("^Im",names(ss2))] <- 1
ss2[grep("^E", names(ss2))] <- 0
ss2[grep("^S",names(ss2))] <- pp[["N"]]/length(aa)-1
sum(ss2)
```

We need to define a cross-age contact matrix. 

* Eventually we should take it from that study (referenced somewhere in the notes) that estimated contact matrices for a variety of countries.
* for now I'm going to define two, one simple "compound symmetric" matrix (1 on the diagonal, 0.1 off-diagonal) and one diagonal. 
* with equal parameters across age categories, what contact matrices will give exactly equal to dynamics to homogeneous mixing? All-equal and diagonal?


```{r Cmats}
aa <- mk_agecats()
## compound symmetric
Cmat <- matrix(0.1, nrow=length(aa), ncol=length(aa), dimnames=list(aa,aa))
diag(Cmat) <- 1
## diagonal
Cmat.d <- diag(length(aa))*length(aa)
dimnames(Cmat.d) <- dimnames(Cmat)
## uniform
Cmat.u <- matrix(1, nrow=length(aa), ncol=length(aa), dimnames=list(aa,aa))
```

At present the indicator for whether the machinery should be run in an
age-structured way or not is the presence/absence of a `Cmat` component in the parameters (now a list
rather than a vector)

Beta "vector" (now a matrix): including only infectious compartments (could also use `full=TRUE`)
```{r beta_vec_etc}
ppa <- c(as.list(pp),list(Cmat=Cmat))
b1 <- make_betavec(ss2, ppa, full=FALSE)
ppa.d <- c(as.list(pp),list(Cmat=Cmat.d))
ppa.u <- c(as.list(pp),list(Cmat=Cmat.u))
b1.d <- make_betavec(ss2, ppa.d, full=FALSE)
plot_grid(ifun(b1,sub="compound"),ifun(b1.d,sub="diag"),nrow=2)
```

Expanded rate matrix: `make_ratemat` bases its categories on the names of the state vector.

```{r make_ratemat}
M <- make_ratemat(ss2, ppa, sparse=TRUE) ## some NA values??
show_ratemat(M)
```

Try `run_sim_range()` [lowest-level/simplest simulation engine]
```{r run_sim_range, warning=FALSE}
rr <- run_sim_range(ppa.u, ss2, nt=200)
par(las=1,bty="l")
matplot(rr[,1],rr[,-1],lty=1,type="l",log="y",xlab="time (days)",ylab="")
```

Try `run_sim`

```{r run_sim}
rr2 <- run_sim(ppa.u, ss2,end_date="2020-10-01",condense=FALSE)
plot(rr2,log=TRUE)+theme(legend.position="none")
```

```{r comp_sims}
pp_list <- list(ppa,ppa.d,pp,ppa.u)
ss_list <- list(ss2,ss2,condense_age(ss2),ss2)
nm <- c("compound","diag","non_age","unif")
sims <- map2(pp_list,ss_list, ~run_sim(.x,.y, end_date="2020-10-01"))
names(sims) <- nm
ks <- c("H","ICU","hosp","death","D")
simplots <- map2(sims,names(sims), ~plot(.x,log=TRUE,log_lwr=1e-3,
                                         keep_states=ks)+ggtitle(.y))
plot_grid(plotlist=simplots,nrow=2)
```

- "compound" example is badly scaled (uniform should just work, and diag example is scaled to keep effective pop sizes the same)
- in these plots we're automatically "condensing" everything, so age categories get aggregated before plotting.  Of course we don't have to (and could write some alternative plot methods that show time X age class grids)
- age-struc runs are missing foi/report/cumRep, because foi is needed to compute report and cumRep and we're not storing foi for age struc right now

Uniform-mixing and non-age are identical (up to small numeric fuzz):

```{r check_equal}
all.equal(droprep(sims$non_age),droprep(sims$unif),tolerance=1e-13)
```

- diag and unif are not *identical* to homogeneous run: visually OK at first glance, but hosp is noticeably different, and closer inspection shows that they're all slightly different

Uniform and diagonal sims match non-age-structured, up to numeric fuzz

```{r plot_diffs}
mm <- mdiff_fun(sims[c("non_age","diag","unif")])
all(na.omit(abs(mm$diff))<1e-8)
```

## age + tests

```{r testify_test}
M.u <- make_ratemat(ss2, ppa.u, sparse=TRUE)
M.ut <- testify(M.u, ppa.u)
show_ratemat(M.ut)
```

## to do

- implement age-dependent N in state construction ...
- try non-homogeneous parameters across age classes; should work to make params a list and replace some of the scalar/length-1 elements with vectors [if params is a list, should test: all elements should be (1) vector of length 1 or (2) vector of length `n_age` or (3) `n_age * n_age` matrix
- foi stuff: time to implement an incidence accumulator and derive foi from it rather than storing foi on the fly? Do we want to store age-structured foi?
- examine age-structured matrix etc ... improve visualization of age/test structured matrices
- clean up make_betavec: do we ever need "full"??
