---
title: "exploring calibration issues"
---

```{r pkgs, message=FALSE}
library(McMasterPandemic)
library(bbmle)
library(dplyr)
library(ggplot2)
library(broom)
library(purrr)
library(dotwhisker)
```


```{r utils}
keep_vars <- c("H","death","report")
kv <- . %>% filter(var %in% keep_vars)
gt <- McMasterPandemic:::get_type
```

Fighting with calibration.
* Nelder-Mead *appears* to converge, but refitting gives a (much) better answer!
* a few earlier trials of BFGS were absolutely awful
* profiling and numeric Hessians (numDeriv::hessian) are very badly behaved (exposes lots of rough edges with how profiling that finds new minima are handled
* decreasing `reltol` doesn't seem to make much difference

```{r}
load("../ontario/ontario_calibration_noICU_2brks_prior.RData")
load("../ontario/ontario_clean.RData")
load("alt_cal.RData")
load("../ontario/ontario_de_cal.RData")
```

Comparing "refit" (restart from fitted parameters)
```{r}
fit_list <- list(orig=fit1,refit=fit2,tol_12=fit3, tol_16=fit4)
m <- map_dbl(fit_list, ~ -logLik(.$mle2))
print(m - min(m))
print(de_cal1$optim$bestval-min(m))  ## 10 log-likelihood units better!
## Nelder-Mead _thinks_ it converged every time!
conv <- map_int(fit_list, ~.$mle2@details$conv)
print(conv)
```

**todo** can we make `slice1D` work to see if these at least *look* like local min? (tried `numDeriv::hessian()`, doesn't do anything nice)

```{r tidy, warning=FALSE} 
tt <- map_dfr(fit_list, ~tidy(.$mle2), .id="fit")
pd <- ggstance::position_dodgev(height=0.25)
print(gg1 <- ggplot(tt,aes(x=estimate,xmin=estimate-2*std.error,
              xmax=estimate+2*std.error,
              y=term,colour=fit))
      + geom_point(position=pd)
      + geom_linerange(position=pd) 
      )
```


```{r pred_tbl}
## extract auxiliary info
f_args <- fit1$forecast_args
ff <- f_args[!names(f_args)=="fixed_pars"]
dd <- ont_all %>% trans_state_vars() %>% kv() %>% gt()
pred_tbl <- (map_dfr(fit_list, ~predict(.) %>% kv() %>% filter(date>min(dd$date)),
                     .id="fit"))
## predict from DE fit
de_pred <- (do.call(forecast_sim, c(list(p = de_cal1$optim$bestmem), f_args)) %>% kv() %>% gt()
    %>% filter(date>min(ont_noICU$date)) %>% mutate(fit="DE")
)
## combine with other preds
pred_tbl <- bind_rows(pred_tbl,de_pred)
## combine with data
pred_tbl2 <- (pred_tbl
    %>% rename(pred="value")
    %>% inner_join(ont_noICU,by=c("date","var"))
    %>% na.omit()
)
```

```{r plot, fig.width=8}
gg2 <- (ggplot(pred_tbl2, aes(date,colour=var))
    + geom_line(aes(lty=fit,y=pred))
    + scale_y_log10()
    + facet_wrap(~var,scale="free")
    + geom_point(aes(y=value))
    + geom_vline(xintercept=fit1$forecast_args$break_dates)
)
print(gg2)
```

Bad Hessian:

```{r bad_hess, cache=TRUE}
vv <- do.call(numDeriv::hessian,
              c(list(func=mle_fun,x=de_cal1$optim$bestmem,data=fit1$mle2@data$data),
                ff))
sqrt(diag(vv))
```

```{r last_pop}
M <- de_cal1$member$pop
colnames(M) <- names(de_cal1$optim$bestmem)
pairs(M,gap=FALSE)
```

```{r pop_lik, cache=TRUE, warning=FALSE}
tmpf <- function(p) do.call(mle_fun,c(list(p,data=fit1$mle2@data$data),ff))
tmpf(de_cal1$optim$bestmem)
nll_vec <- apply(M,1,tmpf)
nll_off <- log10(nll_vec-min(nll_vec))
hist(nll_off)
```

```{r warning=FALSE}
par(las=1)
plot(10^sort(nll_off),log="y",
     ylab="likelihood ratio")
abline(h=10^c(1.5,2),lty=2)
```

```{r}
pairs(M,gap=FALSE,col=ifelse(nll_off<1.5,"black","red"))
```

```{r}
car::scatterplotMatrix(M[nll_off<1.5,],ellipse=TRUE,
                       regLine=FALSE,smooth=FALSE)
```

Can we make a forecasting ensemble based on this?
