---
title: "Getting to know MacPan"
author: "Irena Papst"
output: bookdown::html_notebook2
---

I'm using this doc to record notes as I explore MacPan's codebase. Here's a [handy link](https://jozef.io/r916-exploring-r-code-interactively/) with tools for inspecting code in R (really for debugging).  

```{r load-libraries, echo = FALSE, message = FALSE}
library(McMasterPandemic)
library(ggplot2); theme_set(theme_bw())
library(cowplot)
```

# Setup

## Load params

```{r}
debug(read_params)
params1 <- read_params("ICU1.csv")
print(class(params1))
print(head(params1))
```

Params are loaded in as a named vector, where elements can be accessed as follows:

```{r}
print(params1["alpha"])
```

If a description column (`desc_col`) was provided in the original param file (describing the parameter symbols), it will be stored as the `"description"` attribute (a named list where the parameter symbols are the column names):

```{r}
print(head(attr(params1, "description")))
```

There is also a `describe_params()` method that takes `params_pansim` objects and nice table with both parameter values and meanings aggregated:

```{r}
describe_params(params1)
```

Not sure why each column is of type factor (or if this may cause problems down the road...)

## Summarise params

The `summary` method will take the raw parameters and calculate some derived model quantities:

```{r}
debug(summary)
params1_summary <- summary(params1)
print(params1_summary)
```

