---
title: "Minimal reproducible example: negative jumps"
author: "Irena Papst"
date: "`r format(Sys.time(), '%d %b %Y at %H:%M')`"
output: 
  html_document:
    df_print: "paged"
---

```{r setup, echo = FALSE, message = FALSE}
devtools::load_all()
library(ggplot2)
library(tidyr)
library(dplyr)
```

Here's a minimal example showing that sometimes state variables go negative in a `McMasterPandemic` simulation. This example is reproducible on the `master` branch, so it isn't because of something silly I've done on my `ip_devel` branch.

```{r}
## params
beta0 <- 3
base_params <- update(read_params("PHAC.csv")
                      , beta0 = beta0
                      )
## state
base_state <- make_state(N = 1e6, E0 = 5)

## sim
start_date <- "2020-02-01"
end_date <- "2020-09-01"
res <- run_sim(params = base_params, state = base_state, 
               start_date = start_date, end_date = end_date)

## results
res_long <- (res 
  %>% select(-incidence)
  %>% pivot_longer(-date)
)

neg_states <- (res_long %>% filter(value < 0))

neg_states
```

I'm taking quite a large transmission rate (per capita, per day) of `r beta0`, but it's not an immediate (one time step into the sim) nosedive into a negative state: the sim starts on `r start_date` and the first time step with a negative state variable ($S$) is `r min(neg_states$date)`. 

The time step before the negative jump, the state variables are as follows:

```{r}
day_before_jump <- (res_long
  %>% filter(date == min(neg_states$date)-1) 
)

day_before_jump
```

```{r echo = FALSE}
I_before_jump <- (day_before_jump 
  %>% filter(name == "I") 
  %>% pull(value)
)
```

So at this point we have `r I_before_jump %>% format(scientific = FALSE, big.mark = ",")` infectives, around `r round(I_before_jump/base_params[["N"]]*100)`% of the total population (`r format(base_params[["N"]], scientific = FALSE, big.mark = ",")` people), and each of these infectives is supposed to transmit to `r beta0` susceptibles the next day. This is what causes the (massive) jump into negative susceptibles at the next time step.

The negative jump issue shouldn't be a difficult problem to solve, but we should be smart about how and when a check gets performed to avoid significant slow-downs in the sims. I should also say that this hasn't been a concern for simulations of realistic COVID-19 parameters, but I've come up against it in simulating vaccination for what I thought was a reasonable number of total doses per day (based on Ontario's population size and the actual number of doses per day being administered there currently).


## Plotting

```{r echo = FALSE}
print(ggplot(res_long,aes(x=date,y=value)) 
	+ geom_line() 
	+ facet_wrap(~name,scale="free") 
	+ xlim(c(as.Date("2020-02-01"),as.Date("2020-04-01")))
)
```
