---
title: "analysis of Ontario COVID19 data"
author: "Ben Bolker and Mike Li"
date: "`r format(Sys.time(),'%d %b %Y')`"
---

```{r opts, include=FALSE}
knitr::opts_chunk$set(echo=FALSE)
```

```{r pkgs,message=FALSE}
library(tidyverse)
library(McMasterPandemic)
library(directlabels)
library(ggplot2); theme_set(theme_bw())
library(colorspace)
scale_colour_discrete <- function(...) {
  colorspace::scale_colour_discrete_qualitative(...)
}
scale_fill_discrete <- function(...) {
  colorspace::scale_fill_discrete_qualitative(...)
}
```

Reading data from Mike Li's [curated Canadian COVID data repository](https://github.com/wzmli/COVID19-Canada),
harvested from official sources such as [this](https://www.ontario.ca/page/2019-novel-coronavirus#section-0).

```{r read_data,message=FALSE,warning=FALSE}
url <- "https://raw.githubusercontent.com/wzmli/COVID19-Canada/master/clean.Rout.csv"
## download.file("https://raw.githubusercontent.com/wzmli/COVID19-Canada/master/clean.Rout.csv", dest="today.csv")
dd <- read_csv(url)
## dd %>% filter(Province=="ON") %>% select(Date,source,Note) %>% View()

ont_dd <- (dd
  %>% filter(Province=="ON")
  %>% select(Date,Hospitalization,ICU,ventilator,incidence,deceased)
  %>% rename(pos_tests="incidence")
  %>% pivot_longer(-Date,names_to="var")
  %>% filter(Date>=as.Date("2020-03-15"))
)
```

variable meanings (see https://github.com/wzmli/COVID19-Canada/blob/master/ON.R )

* Hospitalization: Number of patients hospitalized with COVID-19
* ICU: Number of patients in ICU with COVID-19
* Ventilator: Number of patients in ICU on a ventilator with COVID-19
* pos_tests: difference in daily reports of cumulative presumptive + confirmed positive cases
* deceased: cumulative deaths

```{r main_plot,warning=FALSE}
gg0 <- (ggplot(ont_dd,aes(Date,value,colour=var))
  + geom_point()
  + scale_y_log10()
  + geom_line(size=0.5,alpha=0.5)
  + geom_smooth(method="lm",
                formula=y~poly(x,2))
)
dm <- list(dl.trans(x=x+0.2),cex=1,last.bumpup)
print(gg1 <- (gg0
  + geom_dl(aes(label=var),method=dm)
  + theme(legend.position="none")
  + expand_limits(x=as.Date("2020-04-15"))
)
)
```

The curves represent quadratic fits to the data. 

```{r ICU_ratio}
ICU_ratio <- (ont_dd
  %>% filter(var %in% c("ICU","ventilator"))
  %>% pivot_wider(names_from="var",values_from="value")
  %>% na.omit()
  %>% mutate(ratio=ventilator/ICU)
)
```

Source code for this document is [here](https://github.com/bbolker/McMasterPandemic/blob/master/notes/ontario_ICU.Rmd).
