---
title: "analysis of Ontario COVID19 data"
author: "Ben Bolker, Michael Li, Jonathan Dushoff, David Earn (McMaster University)"
date: "`r format(Sys.time(),'%d %b %Y')`"
---

```{r opts, include=FALSE}
knitr::opts_chunk$set(echo=FALSE)
```

```{r pkgs,message=FALSE}
library(tidyverse)
library(glmmTMB)
library(broom.mixed)
library(McMasterPandemic)
library(ggplot2); theme_set(theme_bw())
library(colorspace)
## make nice defaults
scale_colour_discrete <- function(...) {
  colorspace::scale_colour_discrete_qualitative(...)
}
scale_fill_discrete <- function(...) {
  colorspace::scale_fill_discrete_qualitative(...)
}
```

```{r get_data, message=FALSE, warning=FALSE}
source("ontario_clean.R")
```

Reading data from Michael Li's [curated Canadian COVID data repository](https://wzmli.github.io/COVID19-Canada)
harvested from official sources such as [this](https://www.ontario.ca/page/2019-novel-coronavirus#section-0) and [hospitalization cumulative counts](http://wzmli.github.io/COVID19-Canada/PHO.csv) gathered from [Public Health Ontario Epidemiological Summary](https://files.ontario.ca/moh-covid-19-report-en-2020-04-07.pdf). (Latest data in this example are from `r format(max(ont_dd$date))`.)



```{r ONplot,message=FALSE, warning=FALSE}
print(gg1 <- ggplot(ont_dd,aes(date,value,colour=var))
  + geom_point()
  + scale_y_log10()
  + geom_smooth(method="lm",formula=y~poly(x,2))
)
```

- Natural for ICU and Ventilator to be parallel (no real lags here)
- recent nonlinearity/flattening in cases is too recent/sharp to affect the estimated slope much
- Weird that the ICU/vent curves are flattening before the other two: not apparently a capacity issue, maybe something about reporting (e.g. see [here](https://www.ontario.ca/page/2019-novel-coronavirus#7))?

Picture is cleaner (and flattening is more apparent) if we focus on more recent data.

```{r plot_recent, warning=FALSE}
print(gg1 %+% ont_recent)
```

We want to find the first _useful/meaningful_ day for each variable and adjust the different time series to their respective meaningful day.

```{r ont_plot_shift, message=FALSE, warning=FALSE}
print(gg2 <- ggplot(ont_recent,aes(vday,value,colour=var))
    + geom_point()
    + scale_y_log10()
    + geom_smooth(method="lm",formula=y~poly(x,2))
    + scale_x_continuous(limits=c(-1,NA))
)
```

## Calibration strategy

- set reasonable parameters (based on literature; ours come from the Stanford covid-interventions model and from personal communications about health-utilization parameters (fraction hospitalized, length of stays in acute care/ICU, etc.)
- 
- get an estimate of $r(0)$ (initial exponential growth rate)
- (NOT YET) adjust time-varying $r(0)$

## Fitting to data

We want to use the time series data to estimate the growth rate. 

```{r nbfit1, message=FALSE, warning=FALSE}
fit1 <- glmmTMB(value~var-1 + var:vday
  , family=nbinom2
  , dispformula=~var-1
  , data=ont_recent
)
```


Use raw polynomials to parameterize in terms of initial slope rather than average slope.

```{r nbfit_quad, message=FALSE, warning=FALSE}
fit1Q <- update(fit1, . ~ -1 + var + var:poly(vday,2,raw=TRUE))
```

"False convergence" warnings ...


```{r tidying_fits, message=FALSE, warning=FALSE}
fix_term <- function(x) {
  gsub(":poly(vday, 2, raw = TRUE)","Poly"
    , gsub("var","",x)
    , fixed=TRUE
  )
}

t0 <- (tidy(fit1Q,conf.int=TRUE)
  %>% mutate_at("term",fix_term)
  %>% mutate(type=ifelse(grepl("[[:alpha:]]$",term)
          , "int"
          , ifelse(grepl("1$",term),"linear","quad")
          )
      , var=gsub("Poly[0-9]","",term)
      , var=factor(var,levels=rev(
          c("newConfirmations","Hospitalization","ICU","Ventilator","deceased")
          ))
      )
  %>% select(-c(effect,component,statistic,p.value))
)

```

Coefficient plot with tidied frame

```{r coefplot, warning=FALSE, message=FALSE}
print(gg3 <- ggplot(t0,aes(y=var,x=estimate,xmin=conf.low,xmax=conf.high))
    + geom_pointrange()
    + facet_wrap(~type,scale="free",ncol=1)
    + geom_vline(xintercept=0,lty=2)
    + labs(y="")
)
```

- intercepts for variables other than newConfirmations describe sensitivity (how small a non-zero value is actually reported?
- newConfirmations and hosp slopes and curvature agree fairly well
- don't know why death is slower  and ICU/Ventilator are faster (MLi: under reporting? there was a discussion on missing a lot of deaths. BMB: Underreporting alone doesn't change $r$ ... need some time-varying change in reporting rate etc..)
- deaths conf intervals might be overly narrow because these are cumulative values

## calibrate to hospitalization

We have a variety of time series (case reports, hospitalization, ICU, ventilator, death) from which to try to estimate initial growth rate $r_0$ (and potentially any time variation in $r$, which we would usually attribute to behavioural change or physical distancing efforts); we could pick any one of these, or some average.

* in general it is better to fit to incidence-like variables rather than cumulative values or prevalence (e.g. hospital/ICU/ventilation utilization), because cumulative and prevalence values are considerably autocorrelated. We can always make a cumulative variable into an incidence-like variable by differencing it, although in the case of health utilization (hospital, ICU, vent), this will include changes due to both admission and discharge; we would prefer a hospital admission time series if it were available. In general cumulative curves will still give reasonable estimates of the initial growth rate, but they will underestimate uncertainty [REF Rohani and King?]
* case confirmations are difficult to interpret because they are strongly confounded with time-varying testing intensity and criteria.
* when possible, especially when estimating *changes* in $r$ we would prefer to use leading rather than lagging indicators (cases > hospital admissions > hospital occupancy > death)
* in general more severe outcomes are less likely to be underreported or inconsistently reported (deaths > hospitalizations > infections)
* given that health utilization is not too high (saturating/overwhelming resources) ICU, vent, and hospital occupancy are relatively equivalent, given that a reasonably constant fraction of hospitalized cases require ICU/ventilators. (ICU occupancy will probably lag acute-care a bit because the average occupancy time is longer, although there is also a post-ICU acute-care period.)
* we have chosen to calibrate to hospitalizations, although as shown above there is not a big difference between the estimated $r$ based on new confirmations vs. hospital occupancy.

```{r sample_params, message=FALSE,warning=FALSE}
vv <- vcov(fit1Q)$cond
hosp_ind <- grep("Hospitalization",rownames(vv))
pars_0 <- fixef(fit1Q)$cond[hosp_ind]
vv <- vv[hosp_ind, hosp_ind]

## assume for now that Wald approx is OK for sampling distribution
nsim <- 200
set.seed(101)
sim_params <- MASS::mvrnorm(nsim,mu=pars_0,Sigma=vv)
## add Gbar distribution (independent Gaussian sample)
Gbar_prior <- c(mean=6,sd=0.2)
## FIXME: narrow for now, come back and fix calibration issues
##  (does analytic solution work for Gbar???)
sim_params_0 <- c(exp(pars_0[1]),pars_0[2],Gbar=6)
sim_params <- cbind(exp(sim_params[,1]),
                    sim_params[,2],
                    Gbar=do.call(rnorm,c(list(nsim),
                                         as.list(Gbar_prior))))
sim_params <- as.data.frame(sim_params)
names(sim_params) <- names(sim_params_0) <- c("X0","r","Gbar")
```

The first step of calibrating is the get a set of parameters. If we want to compute confidence intervals on the forecasts, we need to generate/sample multiple sets of parameters.

For each set of parameters:

1.  calibrate params to G
2.  calibrate beta0 to r
3.  calibrate intercept
4.  run and store

```{r base_params, warning=FALSE,message=FALSE}
base_params <- read_params(system.file("params","ICU1.csv",
                                       package="McMasterPandemic"))
base_params[["N"]] <- 14.57e6  ## Ontario pop size
reg_date <- (ont_recent
    %>% filter(var=="Hospitalization",vday==0)
    %>% pull(date)
)
## first day of hospitalization data >0/non-NA: 19 March
```

- x= param vector (int/slope/curve/Gbar)
- base_params
- start_date start date for sims
- reg_date start date for regressions
- end_date
- init_var calibration variable
- init_var_value


Calibrate and simulate with point estimates:
```{r calib_point}
s1 <- sim_fun(target=sim_params_0,
      , base_params=base_params
      , reg_date=reg_date
      , end_date="15-Apr-2020"
      , return_val="sim"
        )
## filter(s1,date==ldmy("19-03-2020"))
plot(s1,log=TRUE)+geom_point(data=mutate_at(ont_recent,"var",trans_state_vars))
```

Checking calibration:
```{r calib_check}
tt2 <- (tidy(fit1Q)
    %>% select(term,estimate,std.error)
    %>% filter(grepl("Hosp",term))
    %>% mutate(term=c("int","lin","quad"))
)
ff <- setNames(tt2$estimate,tt2$term)
ffdat <- (tibble(vday=0:20)
    %>% mutate(logpred = ff[["int"]]+vday*ff[["lin"]],
               value=exp(logpred),
               var="Hospitalization"))
(ggplot(ont_recent,aes(vday,value,colour=var))
    + geom_point()
    + scale_y_log10()
    + geom_line(data=ffdat)
)
```


Simulating (calibrating every time; could arguably (more efficiently?) sample over a joint distribution of r, G, E0?  Do importance-sampling weighting?

```{r sim_res, warning=FALSE, message=FALSE,cache=TRUE}
## FIXME: could use pmap ... ?
t1 <- system.time(sim_res <- plyr::alply(sim_params
  , .margins=1
  , .fun=sim_fun
  , base_params=base_params
  , reg_date=reg_date
  , end_date="1-June-2020"
  , return_val="vals_only"
  )
)
```

FIXME: more robust caching/making of this step

Put together pieces.

```{r calibrate, warning=FALSE,message=FALSE}
## put together the pieces (alply gives lists, we want a matrix)
sim_mat <- bind_rows(sim_res)
sim_q <- (t(apply(sim_mat,1 ,quantile,c(0.1,0.5,0.9),na.rm=TRUE))
    %>% as_tibble()
    %>% setNames(c("lwr","median","upr"))
)
## do one simulation to get date/var columns
sim_last <- sim_fun(sim_params[1,], base_params=base_params, reg_date=reg_date,
                 return_val="aggsim",
                 end_date="1-June-2020")
sim_final <- bind_cols(sim_last[c("date","var")], sim_q)
```

## Plotting forecast

```{r forecast,warning=FALSE,message=FALSE}

## FIXME: use trans_var?

label_dat <- data.frame(
  var =c("Hospitalization","ICU","Ventilator","newConfirmations","deceased")
  , sim_var = c("H","ICU","V","report","D")
)

ont_recent2 <- (left_join(ont_recent,label_dat)
    %>% mutate(var=sim_var)
    %>% filter(var!="V") ## not that much point plotting ventilator
)

print(gg_forecast <- ggplot(sim_final,aes(date,median,colour=var))
    + scale_y_log10()
    + geom_line()
    + geom_point(data=ont_recent2,aes(x=date,y=value))
    + geom_ribbon(aes(ymin=lwr,ymax=upr,fill=var), colour=NA,
                  alpha=0.2)
    + geom_vline(xintercept=reg_date,lty=2)
    + facet_wrap(~var)
)
```

- weird patterns: because of calibration of intercept??? looks plausible.
- implied curves of R_t?
- dates of control???

## estimating breakpoints

```{r breakpoint}
pp <- calib2(base_params=base_params,
             target=sim_params_0,
             start_date="1-Mar-2020",
             reg_date="15-Mar-2020",
             init_var="H")
b <- get_break_old(params=pp,data=ont_recent)
print(b$par)  ## logit(rel beta0), log(nb_disp)
print(c(plogis(b$par[1]),exp(b$par[2])))
bb <- run_sim_break(params=pp,rel_beta0=plogis(b$par[1]),break_dates=ldmy("20-Mar-2020"),
                    start_date="1-Mar-2020")
plot(bb,log=TRUE)+  geom_point(data=ont_recent2,aes(x=date,y=value))
```

```{r plot_long_sim}
bb_long <- run_sim_break(params=pp,rel_beta0=plogis(b$par[1]),break_dates=ldmy("20-Mar-2020"),
                    start_date="1-Mar-2020",end_date="1-Aug-2020")
(plot(bb_long,log=TRUE)
    + geom_point(data=ont_recent2,aes(x=date,y=value))
    + facet_wrap(~var)
    + theme(legend.position="none")
)
```

Summarize R0 etc. before and after breakpoint

What's up with cases???

## combined calibration

What if we want to estimate `beta0`, `E0`, and more than one breakpoint at the same time? (Primary use case here is for South African data where the 

## next

1. better/different approaches to estimating time-varying beta?




