---
title: "analysis of Ontario COVID19 data"
author: "Ben Bolker, Michael Li, Jonathan Dushoff, David Earn (McMaster University)"
date: "`r format(Sys.time(),'%d %b %Y')`"
---

```{r opts, include=FALSE}
knitr::opts_chunk$set(echo=FALSE)
```

```{r pkgs,message=FALSE}
library(tidyverse)
library(glmmTMB)
library(nlme)
library(broom.mixed)
library(McMasterPandemic)
```


Reading data from Michael Li's [curated Canadian COVID data repository](https://wzmli.github.io/COVID19-Canada)
harvested from official sources such as [this](https://www.ontario.ca/page/2019-novel-coronavirus#section-0) and an [hospitalization cumulative counts](http://wzmli.github.io/COVID19-Canada/PHO.csv) gather from Public Health Ontario situation reports.


```{r read_data,message=FALSE,warning=FALSE}
url <- "https://wzmli.github.io/COVID19-Canada/git_push/clean.Rout.csv"
dd <- read_csv(url)

PHOurl <- "http://wzmli.github.io/COVID19-Canada/PHO.csv"
ddPHO <- read_csv(PHOurl)

```

```{r data cleaning,message=FALSE,warning=FALSE}

ont_dd <- (dd
  %>% filter(Province=="ON")
  %>% select(Date,Hospitalization,ICU,Ventilator,deceased,newConfirmations)
  %>% pivot_longer(-Date,names_to="var")
  %>% filter(Date>=as.Date("2020-03-15"))
)
```


##'  Filter/date by time since first reported case, per province ...
```

##' Warm up with a quadratic fit (changing scale in ggplot link function in GLMs) to reported incidence in all provinces
##+ first_plot
print(gg0 <- ggplot(dd,aes(Date,newConfirmations,colour=Province))
    + geom_line(size=0.5,alpha=0.3)
    + geom_point()+scale_y_log10()
    + geom_smooth(method="lm",
                  formula=y~poly(x,2),
                  se=FALSE)
)

## easiest way to plot mixed model outputs with CIs?
## sjPlot, broom.mixed, ... ?

## data exploration and possible comparison of IDEA/IHME/etc. approaches
## (i.e. purely phenomenological)
## fit linear, quad mixed? models
## logit, Richards?
## epigrowthfit or mle2 or ... ?
## plot predictions and CIs

ont_dd <- (dd
    %>% filter(Province=="ON")
    %>% select(Date,Hospitalization,ICU,Ventilator,newConfirmations,deceased)
    %>% pivot_longer(-Date,names_to="var")
)

print(gg1 <- ggplot(ont_dd,aes(Date,value,colour=var))
    + geom_point()
    + scale_y_log10()
    + geom_smooth(method="lm",
                  formula=y~poly(x,2))
)
##' * Natural for ICU and Ventilator to be parallel (no real lags here)
##' * recent nonlinearity/flattening in cases is too recent/sharp to affect the estimated slope much
##' * very weird that the ICU/vent curves are flattening before the other two

##' Picture is cleaner (and flattening is more apparent) if we focus on more recent data:
##'

## find first useful day
min_day <- function(day,value) {
    good <- !is.na(value) & value>0
    if (all(good)) min(day) else min(day[good])
}

ont_recent <- (filter(ont_dd,Date>=as.Date("2020-03-15"))
    %>% mutate(day=as.numeric(Date-min(Date)))
    %>% group_by(var)
    %>% mutate(vday=day-min_day(day,value))
    %>% ungroup()
)
    
print(gg1 %+% ont_recent)

print(gg2 <- ggplot(ont_recent,aes(vday,value,colour=var))
    + geom_point()
    + scale_y_log10()
    + geom_smooth(method="lm",
                  formula=y~poly(x,2))
    + scale_x_continuous(limits=c(-1,NA))
)

fit1 <- glmmTMB(value~var-1 + var:vday,
        family=nbinom2,
        dispformula=~var-1,
        data=ont_recent)

## FIXME: check vs nbinom1 ?

##' use raw polynomials to parameterize in terms of initial slope
##' rather than average slope
fit1Q <- update(fit1, . ~ -1 + var + var:poly(vday,2,raw=TRUE))
## false convergence warning?

fix_term <- function(x) {
    gsub(":poly(vday, 2, raw = TRUE)","Poly",
         gsub("var","",x),
         fixed=TRUE)
}

t0 <- (tidy(fit1Q,conf.int=TRUE)
    %>% mutate_at("term",fix_term)
    %>% mutate(type=ifelse(grepl("[[:alpha:]]$",term),"int",
                       ifelse(grepl("1$",term),"linear","quad")))
    %>% mutate(var=gsub("Poly[0-9]","",term),
               ## reverse levels to get top-to-bottom
               var=factor(var,levels=rev(c("newConfirmations","Hospitalization",
                                       "ICU","Ventilator","deceased"))))
    %>% select(-c(effect,component,statistic,p.value))
)
print(gg3 <- ggplot(t0,aes(y=var,x=estimate,xmin=conf.low,xmax=conf.high))
    + geom_pointrange()
    + facet_wrap(~type,scale="free",ncol=1)
    + geom_vline(xintercept=0,lty=2)
)

##' * intercepts for variables other than newConfirmations describe sensitivity (how small a non-zero value is actually reported?
##' * newConfirmations and hosp slopes and curvature agree fairly well
##' * no 'clear' evidence of slowing newConfirmations/hospitalization (¿ flattening in newConfirmations could be obscured by increasing testing/ too recent or sharp to be well-modeled by a quadratic ?)
##' * 
##' * don't know why death is slower  and ICU/Ventilator are faster
##' * MLi: maybe underreporting in death
##' * deaths conf intervals might be overly narrow because these are cumulative values

##' calibrate to hospitalization

vv <- vcov(fit1Q)$cond
hosp_ind <- grep("Hospitalization",rownames(vv))
mu <- fixef(fit1Q)$cond[hosp_ind]
vv <- vv[hosp_ind, hosp_ind]

## assume for now that Wald approx is OK for sampling distribution
nsim <- 200
set.seed(101)
sim_params <- MASS::mvrnorm(nsim,mu=mu,Sigma=vv)
## add Gbar distribution (independent Gaussian sample)
Gbar_prior <- c(mean=6,sd=0.2)
## FIXME: narrow for now, come back and fix calibration issues
##  (does analytic solution work for Gbar???)
sim_params <- cbind(sim_params,Gbar=do.call(rnorm,c(list(nsim),
                                                    as.list(Gbar_prior))))
sim_params <- as.data.frame(sim_params)
names(sim_params) <- c("int","slope","curve","Gbar")

## for each set of parameters:
## (1) calibrate params to G
## (2) calibrate beta0 to r
## (3) calibrate intercept
## (4) run and store


base_params <- read_params(system.file("params","ICU1.csv",
                                       package="McMasterPandemic"))
base_params[["N"]] <- 14.57e6  ## Ontario pop size
reg_date <- (ont_recent
    %>% filter(var=="Hospitalization",vday==0)
    %>% pull(Date)
)

##' x= param vector (int/slope/curve/Gbar)
##' base_params
##' start_date start date for sims
##' reg_date start date for regressions
##' end_date
##' init_var calibration variable
##' init_var_value
sim_fun <- function(x, base_params,
                    start_date="1-Mar-2020",
                    reg_date,
                    end_date="1-May-2020",
                    init_var="H", sim_args=NULL,
                    vals_only=TRUE) {
    ## calibration
    p <- fix_pars(base_params, target=c(Gbar=x[["Gbar"]]),
                  pars_adj=list(c("sigma","gamma_s","gamma_m","gamma_a")))
    stopifnot(all.equal(get_Gbar(p),x[["Gbar"]],tolerance=1e-4))
    ## fix params to *initial* slope (for now)
    p <- fix_pars(p, target=c(r=x[["slope"]]),
                  pars_adj=list("beta0"),u_interval=c(-1,2))
    ## FIXME: problems with r-calibration here if using r_method="rmult"
    stopifnot(all.equal(get_r(p),x[["slope"]],tolerance=1e-4))
    ## calibrate initial state (shooting)
    state <- get_init(date0=start_date,date1=reg_date,p,var=init_var,
                      init_target=x[["int"]], sim_args=sim_args)
    arg_list <- nlist(params=p,
                      state,
                      start_date,
                      end_date)
    r <- do.call(run_sim,arg_list)
    a <- aggregate(r,pivot=TRUE)
    if (vals_only) return(a$value) else return(a)
}

## FIXME: could use pmap ... ?
t1 <- system.time(sim_res <- plyr::alply(sim_params, .margins=1, .fun=sim_fun,
                       base_params=base_params, reg_date=reg_date,
                       end_date="1-June-2020",
                       .progress="text"))

## put together the pieces (alply gives lists, we want a matrix)
sim_mat <- bind_rows(sim_res)
sim_q <- (t(apply(sim_mat,1 ,quantile,c(0.1,0.5,0.9),na.rm=TRUE))
    %>% as_tibble()
    %>% setNames(c("lwr","median","upr"))
)
sim_0 <- sim_fun(sim_params[1,], base_params=base_params, reg_date=reg_date, vals_only=FALSE,
                       end_date="1-June-2020")
sim_final <- bind_cols(sim_0[c("date","var")], sim_q)
(gg_forecast <- ggplot(sim_final,aes(date,median,colour=var))
    + scale_y_log10()
    + geom_line()
    + geom_ribbon(aes(ymin=lwr,ymax=upr,fill=var), colour=NA,
                  alpha=0.2)
    + geom_vline(xintercept=reg_date,lty=2)
)

ggsave(gg_forecast,file="forecast0.pdf")
## weird patterns: because of calibration of intercept???
## looks plausible.


## implied curves of R_t?
## dates of control???

### ISSUES


## 1. non-constant r (use epigrowthfit and feed in observed relative r over time (i.e. derivative of fitted curve) as timevars?
## 2. non-constant testing (ignore this for now, maybe we can do a brute-force correction where we subtract r(testing) from r(reports) - this assumes that tests are not expanding to incorporate a different subset of people
## 3. use H, ICU, D to get r then use reports to get time-varying beta0?

