---
title: "visualizing rate matrices"
---

Let's examine different ways to compute and visualize rate matrices.

```{r pkgs, message=FALSE}
library(McMasterPandemic)
library(igraph)
library(Matrix)
library(diagram)
library(latticeExtra) ## layer()
library(lattice)
```

Regular rate matrix.

```{r}
pp <- read_params("PHAC_testify.csv")
state1 <- make_state(params=pp)   ## construct state vector
state1[] <- 1  ## occupy all states so all possible arrows are >0
M1 <- make_ratemat(state1,pp)
M1[M1>0] <- 1 ## set all edges==1 (edge scaling etc.)
g <- igraph::graph_from_adjacency_matrix(M1)
plot(g, layout=igraph::layout_as_tree)
```

This looks OK, except for the fact that the layout is *slightly* funky; with `layout_as_tree` the positioning depends only on the number of steps from the source (`S`). More complex layouts could get messier, and there's not really any automatic layout algorithm that works satisfactorily (I've also tried `layout_nicely` and `layout_with_graphopt`.)

As matrix.  `heatmap` (with `Rowv=NA`, `Colv=NA`, `scale="none"`) is probably the quickest way to get a labelled matrix, but the `image` method from `Matrix` is probably slightly better.

```{r}
## ?Matrix::image-methods
mfun <- function(M, aspect="iso") {
    Matrix::image(Matrix(M), scales=list(x=list(at=seq(nrow(M)),labels=rownames(M)),
                                 y=list(at=seq(ncol(M)),labels=colnames(M))),
                  xlab="to",
                  ylab="from",
                  sub="",
                  aspect=aspect)
}
mfun(M1)
```

As diagram. This isn't right, and definitely needs more tweaking, but is in principle the prettiest version.  (We can also change the adjacency matrix to include symbolic definitions of *per capita* flows.)
(Why aren't H, ICUs, ICUd lined up vertically ??? Does plotmat force positions to let arrows work??)

```{r}
xpos <- c(S=1,E=2,Ia=3,Ip=3,Im=4,Is=4,H=5,ICUs=5,ICUd=5,H2=6,D=7,R=7,X=7)
ypos <- c(S=1,E=1,Ia=1,Ip=2,Im=2,Is=3,H=3,ICUs=4,ICUd=5,H2=4,D=5,R=1,X=4)
pos <- cbind(xpos,ypos)/8
M1d <- M1[names(xpos),names(xpos)] ## reorder ... does that matter?
diagram::plotmat(t(M1d),pos=pos,name=colnames(M1),box.size=0.02, add=FALSE)
```

## test-based version

```{r}
Mt <- testify(make_ratemat(state1,pp), pp)
Mt[Mt>0] <- 1
```

We can plot with igraph, but we can't see anything ...

```{r}
gt <- igraph::graph_from_adjacency_matrix(Mt)
plot(gt, layout=igraph::layout_with_graphopt)
```

Much better if we use a matrix. (Stretch aspect ratio a little to make more room for labels.)

```{r fig.width=20,fig.height=10}
mfun(Mt,aspect=0.5)
```

Checking a little more carefully by zooming in:

```{r fig.width=10,fig.height=10}
w <- which(rownames(Mt)=="Is_t")
(mfun(Mt[1:w,1:w])
    ## lines delineating blocks
    + layer(panel.abline(h=4.5+seq(0,20,by=4),col=2))
    + layer(panel.abline(v=4.5+seq(0,20,by=4),col=2))
)
```

We could do this with Kronecker products!

```{r fig.width=10,fig.height=10}
(mfun(Mt[(w+1):nrow(Mt),(w+1):nrow(Mt)])
    ## lines delineating blocks
    + layer(panel.abline(h=4.5+seq(0,28,by=4),col=2))
    + layer(panel.abline(v=4.5+seq(0,28,by=4),col=2))
)
```

We can see that this is not quite working yet because `D`, `R`, and `X` compartments are not receiving input!


