---
title: refactoring
---

How do we go about modifying/refactoring the code to allow it to more flexibly/generally handle different models of time-varying parameters? 

## issues

- Clarity and maintainability. The more layers we have, the harder it is to figure out how everything fits together
- Performance. Things that will slow down runs (we don't care that much, except that we often run lots of sims for ensembles, estimation, etc.):
    - adding lots of compartments (e.g. Erlang-ization, age structure, spatial structure)
	- many layers of function calls, e.g. calling `run_sim_range` for every day?
- maintaining possibility to recode for speed/latent-ization (pomp, odin, Rcpp, TMB ...)
    - avoid date manipulations in innermost loop
	- avoid character/factor processing (e.g. use `enum`-based switches if necessary)
- how do we pass information/parameters along the chain; via `...` or `do.call()` (or stored info in objects)?
- storing metadata (as attribute, or list element, or ...)
    - attributes leave the behaviour of the original object unchanged
	- list elements are a little more transparent, but may add another layer of extraction
	- accessor/extractor methods??
	- don't even talk about S4
- post-processing of state variables
    - condensation (i.e. collapsing multiple state variables to one)
	- differencing
	- cumulating
- processing of parameters
    - unlisting/relisting
	- linking/inverse-linking

## possibilities

- I'm thinking about generalizing `run_sim_break` by allowing other functions to specify the time variation of parameters (see `OTHER_TIMEFUN` below)
- collapse some of `run_sim`/`run_sim_range`/`do_step`?  Calling `run_sim_range` for every step, as we would have to do with more continuously time-varying parameters, seems awkward. On the other hand, `run_sim_range` was created in the first place so that we didn't have to deal with date-processing at the lowest level.  Maybe `run_sim` could process dates into a zero-based numeric format/find numeric offsets for parameters, so that `run_sim_range` was continuous but low-level (i.e. suitable for coding in TMB/Stan/etc.)?
- collapse `forecast_sim`/`run_sim_break`? This made more sense before I started thinking about adding tracks parallel to `run_sim_break` ...

```{r plot_graph,message=FALSE,echo=FALSE,fig.width=8,fig.height=8}
library(mvbutils)
library(igraph)
library(McMasterPandemic)
pos <- which(search()=="package:McMasterPandemic")
## FIXME: include predict methods?
funs <- grep("mle_fun|forecast|calibrate$|run_|do_step",ls(pos=pos),value=TRUE)
funs <- c(funs,"predict.fit_pansim")
ff <- foodweb(where=pos,
              funs=funs,
              ## rprune="forecast|calibrate|run_",
              plotting=FALSE, ancestors=FALSE,descendents=FALSE)
## HACK: foodweb doesn't recognize do.call() or ... ?
M <- ff$funmat
M["predict.fit_pansim",c("forecast_sim","forecast_ensemble")] <- 1
## HACK: set up another function parallel to run_sim_break
## M <- rbind(M,0)
## M <- cbind(M,0)
## rownames(M)[nrow(M)] <- "OTHER_TIMEFUN"
## colnames(M)[ncol(M)] <- "OTHER_TIMEFUN"
## M[,"OTHER_TIMEFUN"] <- M[,"run_sim_break"]
## M["OTHER_TIMEFUN",] <- M["run_sim_break",]
## HACK: calibrate effectively calls the other run_sim
run_sim_funs <- setdiff(grep("run_sim_",rownames(M),value=TRUE),"run_sim_range")
M["forecast_sim",run_sim_funs] <- 1
M[run_sim_funs,"run_sim"] <- 1
g <- igraph::graph_from_adjacency_matrix(M)
plot(g,layout=layout.graphopt) ## fruchterman.reingold
```

## parameter passing

How do we deal efficiently and transparently with parameters that need to get passed/used in different places? At the moment the structure (and starting values) is stored in `opt_pars`, but different parts are used by different components. We want to save as much as necessary and pass the right pieces ...

- `params` gets used only (?) within `run_sim()` and `run_sim_range()` (and in various summary methods, Jacobian calcs, etc.)
- `(log_)nb_disp` gets only used in `mle_fun`; it NULLs any elements containing `nb_disp` before passing
- `time_args` includes components

## Functions

### `calibrate`

```{r echo=FALSE}
show_args <- function(f) {
    p <- paste(names(formals(f)), collapse=", ")
    cat("**parameters:** ",strwrap(p),sep="\n")
}
```

takes data and a set of starting parameters/structure/etc. and uses `DEoptim` and `mle2` to estimate parameters by trajectory matching

```{r echo=FALSE,results="asis"}
show_args(calibrate)
```

- `sim_args` is passed down, eventually to `run_sim`. Can we also use it for arguments to `sim_fun` (the one with time-varying stuff) and strip those arguments before passing it down - or use `...` to catch junk in `run_sim`?

### `mle_fun`

- takes data and a set of starting parameters/structure/etc. and returns a negative log-likelihood (or a log-posterior-probability)
- it's useful to have this defined separately from/outside `calibrate` so that we can do other things with the log-likelihood (e.g. calculate importance weights for ensembles)



```{r echo=FALSE,results="asis"}
show_args(mle_fun)
```

### `forecast_sim`

inverse-link transforms parameters, re-lists them into the format of `opt_pars`, then calls `run_sim_break`; then condenses/aggregates/pivots results to match data format 

```{r echo=FALSE,results="asis"}
show_args(forecast_sim)
```
`p` (numeric, named parameter vector); `opt_pars` (list: starting values *and structure* for "re-listing")

### `forecast_ensemble`

Calls `forecast_sim` repeatedly to generate an ensemble of trajectories, then computes (possibly weighted) quantiles (alternately may return an array of (replicate $\times$ time $\times$ variable).  May also (indirectly) call `mle_fun` if importance weights are requested.

```{r echo=FALSE,results="asis"}
show_args(forecast_ensemble)
```


### `run_sim_break`


Thin wrapper for `run_sim`: converts `break_dates` plus relative beta information (`rel_beta0`) into a data frame of points at which parameters change, then passes that info plus parameters to `run_sim`. `OTHER_TIMEFUN` would set up different time-dependence in parameters based on dates, parameters (and other covariates such as mobility?)

```{r echo=FALSE,results="asis"}
show_args(run_sim_break)
```
 
### `run_sim`

Constructs rate matrix; processes information about when to turn on stochasticity. Loops over change points, calling `run_sim_range` repeatedly. Currently assumes that only foi changes

```{r echo=FALSE,results="asis"}
show_args(run_sim)
```


### `run_sim_range`

Run multiple simulation steps. Assumes constant *per capita* rates (except for foi, which is updated at each step)


```{r echo=FALSE,results="asis"}
show_args(run_sim_range)
```

### `do_step`

Run a single simulation step. Updates foi (redundantly?); generates total flows/derivatives from rate matrix and applies them

```{r echo=FALSE,results="asis"}
show_args(do_step)
```

