---
title: "selection coefficients"
---

Based on math and common sense and what's going on in MacPan, what are the relationships
between $r$, $R_0$, and `beta0` ? (And what exactly do we mean by "selection coefficient"?)

For a simple SIR model, we have $r = \beta - \gamma$, and the SC is (I think) is the
difference in relative fitness, i.e. $(r_m-r_w)/r_w = r_m/r_w - 1$ ($m$=mutant, $w$=wild type),
so a change in $\beta$ only of $\Delta \beta$ will give a selection coefficient of

$$
\frac{\beta+\Delta beta - \gamma}{\beta - \gamma} - 1 =
1 + \frac{\Delta beta}{r} -1 = \frac{\Delta beta}{r}
$$

```{r}
library(McMasterPandemic)
p1 <- read_params("ICU1.csv")
p2 <- update(p1, beta0=p1[["beta0"]]*1.1)
print(get_r(p2)/get_r(p1)-1) ## proportional change in r
print(0.1/get_r(p1))  ## Delta-beta0/r
```

Compute
```{r}
new_r <- function(m) get_r(update(p1,beta0=p1[["beta0"]]*(1+m)))
bvec <- seq(0,1.5,length=51)
rvec <- sapply(bvec, new_r)
plot(bvec,rvec,xlab="proportional change in beta", ylab="r",type="l")
```

To get from parameters to SC, use `get_r()` and compare to original.
To specify SC, use `fix_pars()` ...

## Proportional change in beta vs difference in r

```{r}
rdiffvec <- rvec - min(rvec)
plot(bvec,rdiffvec,xlab="proportional change in beta", ylab="rdiff",type="l")
m1 <- lm(rvec~bvec) ## can we say why change in r is ~ 0.15*delta_beta?
coef(m1)
abline(m1,col="red")
p3 <- fix_pars(target=c(r=get_r(p1)*1.1),
         params=p1,
         u_interval=c(0,0.5), ## interval for adjusting values:
         ## range from 1*original to exp(0.5)*original
         ## multiply original value by {exp(lwr)} up to {exp(upr)}
         pars_adj="beta0")
get_r(p3)/get_r(p1)
```
