---
output: html_document
---

# Dependence of State Transition Rates on Parameters and State Variables

Currently MacPan developers specify new models by writing R code that transforms parameters and state variables into state transition rates (`make_ratemat`, `update_ratemat`). As we build more of MacPan in `C++` we would like to maintain the ability of epidemiologically-focused developers to define new models in R as much as possible. At the same time we would like to have `C++` developers focused on computational efficiency rather than details of epidemiological models. In order to achieve this separation of concerns, we will need to formalize a spec for a general model of how state transition rates depend on parameters and state variables. This spec will make it easier for the two types of developers to work together. We will version this spec so that we can build it in steps, making sure that we don't over-engineer it. 

This is a working document that provides background on current thinking, spec definitions (currently only drafts), and a roadmap for planning and remembering details that are not yet sorted out. Over time items from the roadmap will move into defined spec versions. To update it use `make` from this folder.

## Versioned Rate Matrix Models

There are currently no versioned rate matrix model specs. See below for experimental specs.

## Experimental Rate Matrix Models

### Warm Up Rate Matrix Model

Allow each element of the rate matrix to depend on products of parameters (or their complements).

Let $M = [M_{i,j}]$ be the rate matrix and $\theta = [\theta_k]$ be the parameter vector. The elements of $M$ are functions of the elements of $\theta$. The dependence of $M$ on $\theta$ can only take certain forms. In particular, each element of $M$ is a product of zero or more elements of $\theta$ or their complements (i.e. $1-\theta$). Here are some examples:

* $M_{4,5} = \theta_1 \theta_2$
* $M_{1,2} = (1-\theta_3)\theta_4\theta_5(1-\theta_{10})$
* $M_{10,3} = (1-\theta_4)$

In general,

$M_{i,j} = \prod_k \theta_k^{x_{ijk}} (1-\theta_k)^{y_{ijk}}$

where $x_{ijk}$ (or $y_{ijk}$) is one if $M_{i,j}$ depends on $\theta_k$ (or $1 - \theta_k$) and zero otherwise. Note that these $x$ and $y$ numbers define the dependence of $M$ on $\theta$, and are constants that do not change throughout a simulation.

This model allows one to compute simple rate matrix structure where the elements of the rate matrix are simply parameters.  An example in MacPan is the following.
```{r, eval=FALSE}
afun("Ia", "R", gamma_a)
```
This model also allows products of parameters and complements of parameters, such as this example from MacPan.
```{r, eval=FALSE}
afun("Is", "ICUs", (1 - nonhosp_mort) * (1 - phi1) * (1 - phi2) * gamma_s)
```

### More General Rate Matrix Model

#### Motivation for generalization

The above warm up model is too restrictive for some rate matrix elements, with the most important example being the force of infection. Here is a definition of the force of infection that is equivalent (I think) to what MacPan uses (at least in some cases).
```{r, eval=FALSE}
afun("S", "E", sum(state[c("Ia", "Ip", "Im", "Is")] 
  * beta0 
  * c(Ca, Cp, (1 - iso_m) * Cm, (1 - iso_s) * Cs) 
  / N))
```

To accommodate such force of infection updates and others, we define a more general model that allows the following additional operations.

1. state variables can be used, in addition to parameters
2. parameters and state variables can be inverted, in addition to complements
3. products of parameters, state variables, and their inverses and complements can be added together

#### User Interface

Users can define the structure of a rate matrix with a list of expressions, each determining the parameter and state dependence of a non-zero rate matrix element. For example,
```{r, eval=FALSE}
mk_ratemat_struct(
  # recovery
  rate(from = "Ia", to = "R",    
       formula = ~ (gamma_a)),
  # hospitalizations
  rate(from = "Is", to = "ICUs", 
       formula = ~ (1 - nonhosp_mort) * (1 - phi1) * (1 - phi2) * (gamma_s)),
  # force of infection
  rate(from = "S",  to = "E",
       formula = ~ 
         (Ia) * (beta0) * (1/N) * (Ca) + 
         (Ip) * (beta0) * (1/N) * (Cp) + 
         (Im) * (beta0) * (1/N) * (Cm) * (1-iso_m) + 
         (Is) * (beta0) * (1/N) * (Cs) * (1-iso_m)),
  rate('etc...')
)
```
The formulas allow the following operations:

* Any element, $x$, of either the parameter or state vector can be placed _in parentheses_ to produce a _factor_ in one of the following three forms
  * Identity: `(x)`
  * Complement: `(1-x)`
  * Inverse: `(1/x)`
* Any number of factors can be multiplied together using `*` to produce a _product_
* Any number of factors and products can be added together using `+`

#### Example

The `utilities.R` file in the folder for this document contains some prototype functions for this approach. Using them we illustrate how rate matrix structure can be flexibly described by the user.
```{r}
source('utilities.R')
rs = mk_ratemat_struct(
  rate("E", "Ia", ~ (alpha) * (sigma)),
  rate("E", "Ip", ~ (1 - alpha) * (sigma)),
  rate("Ia", "R", ~ (gamma_a)),
  rate("Ip", "Im", ~ (mu) * (gamma_p)),
  rate("Ip", "Is", ~ (1 - mu) * (gamma_p)),
  rate("Im", "R", ~ (gamma_m)),
  rate("Is", "H", ~ (1 - nonhosp_mort) * (phi1) * (gamma_s)),
  rate("Is", "ICUs", ~ (1 - nonhosp_mort) * (1 - phi1) * (1 - phi2) * (gamma_s)),
  rate("Is", "ICUd", ~ (1 - nonhosp_mort) * (1 - phi1) * (phi2) * (gamma_s)),
  rate("Is", "D", ~ (nonhosp_mort) * (gamma_s)),
  rate("ICUs", "H2", ~ (psi1)), ## ICU to post-ICU acute care
  rate("ICUd", "D", ~ (psi2)), ## ICU to death
  rate("H2", "R", ~ (psi3)), ## post-ICU to discharge
  ## H now means 'acute care' only; all H survive & are discharged
  #list(from = "H", to = "D", formula = ~ 0),
  rate("H", "R", ~ (rho)), ## all acute-care survive
  rate("Is", "X", ~ (1 - nonhosp_mort) * (phi1) * (gamma_s)), ## assuming that hosp admissions mean *all* (acute-care + ICU)
  # force of infection
  rate("S",  "E", ~
         (Ia) * (beta0) * (1/N) * (Ca) +
         (Ip) * (beta0) * (1/N) * (Cp) +
         (Im) * (beta0) * (1/N) * (Cm) * (1-iso_m) +
         (Is) * (beta0) * (1/N) * (Cs) * (1-iso_m))
)

# sanity check:
all.equal(
  c(do_step2(state, M, params, rs)),
  c(do_step(state, params, M, do_hazard = FALSE, do_exponential = FALSE))
)
```

We can parse this structure to produce information about how to update the elements of the rate matrix. First, all of the _factors_ over all rate matrix entries in this model can be displayed.
```{r}
attributes(rs)$all_factors
```

Here is an example where this input,
```{r}
rs$Ip_to_Is[c('from', 'to', 'formula')]
```
parses as this,
```{r}
rs$Ip_to_Is$factors
```
More generally, all elements of a `ratemat-struct` object are `rate-struct` objects and contain the following elements:
```{r}
names(rs$Is_to_H)
```

The `from` and `to` elements give the name of the associated states, `formula` gives the expression determining the rate, `factors` gives a data frame where each row is a _factor_ (as defined above) required to compute the rate, and `ratemat_indices` gives the row and column index of the rate in the rate matrix.

#### Mathematical Model Description

We make a series of transformations from the state vector, $s$, and parameter vector to the rate matrix, $M$. Given how TMB works, we need to distinguish parameters that will be optimized, $\theta$, from those that will remain constant, $\eta$ (or those that are derived from parameters being optimized, e.g. time-varying??, still unsure how time varying machinery is working with run_sim_break or will work with run_sim_loglin). We define two intermediate vectors as well: the factor vector, $v$, and the product vector, $u$. At a high level we take the elements of the state vector and two parameter vectors into the non-zero elements of the rate matrix as follows.
$$
\{s,\eta,\theta\} \rightarrow v \rightarrow u \rightarrow M
$$

The factor vector, $v$, is a function of $s$, $\eta$, and $\theta$.  The dependence is simple in that each scalar element of the factor vector can be one of the following.

* [identity] an element, $x$, of $s$, $\eta$, or $\theta$
* [complement] the complement of this element, $1-x$
* [inverse] the inverse of this element, $1/x$

In the future we can generalize this by adding more operations, but for now identity, complement, and inverse should get us pretty far -- although it will make age structure awkward given that such problems are more naturally handled with matrix operations like Kronecker products and sweeps.

The product vector, $u$, is a function of $v$.  The dependence is simple in that each element of the product vector is the product of one or more elements of the factor vector, $v$.

The non-zero elements of the rate matrix, $M$, can be any one of the following.

* An element of the factor vector, $v$
* An element of the product vector, $u$
* The sum of one or more elements in the product vector, $u$

Explicitly expressing the force of infection in terms of this model we have.

$$
s = [I_a, I_p, I_m, I_s]
$$

$$
\theta = [\beta_0, C_a, C_p, C_m, C_s, iso_m, iso_s, N]
$$

$$
u = [s_1, s_2, s_3, s_4, \theta_1, \theta_2, \theta_3, \theta_4, \theta_5, 1 - \theta_6, 1 - \theta_7, 1/\theta_8]
$$

$$
v = [u_1  u_5  u_6          u_{12}, 
     u_2  u_5  u_7          u_{12},
     u_3  u_5  u_{10}  u_8  u_{12},
     u_4  u_5  u_{11}  u_9  u_{12}]
$$

And the non-zero element of $M$ determining the rate of flow from S to E is the sum of the elements in the products vector, $v$. Typically this summation will be taken over a subset of the elements of the products vector, but we simplified this example to include on elements that are involved in the force of infection computation.

## Roadmap

### Avoid Copying on the R-Side

Although the overall motivation for this work is to facilitate C++/TMB refactoring, it would be good to have the R-side version as efficient as possible.

### Constants

Could be convenient to just allow users to hard code constant rate matrix entries in the rate matrix structure. This is an example from MacPan.
```{r, eval=FALSE}
afun("H", "D", 0)
```
I'm surprised this is necessary though because I have been assuming that unspecified rate matrix elements imply zero.

### Regex Matching

Not thinking about this at all, but probably should.  For example, could `pfun` return multiple rate matrix positions?  If so that would screw up the current index machinery.

### Parameters to optimize vs other parameters

Because of how TMB is used to produce objective functions for optimization, the parameters to optimize/calibrate must be separated from the other parameters. Currently I have ignored this complexity, but I think we don't need to think about it until we actually implement this on the `C++` side. This separation of parameters will still need to be done thoughtfully though, because it might have implications for accessing memory efficiently during simulation steps.

### Index Permutations

Permutation of the parameter, state, and factor vectors for computational efficiency. We don't want to have to sum together products that depend on elements of the state vector that are far apart from each other in memory. This is similar to permutations of sparse matrices in the Matrix package and Eigen.

Made some progress here -- we can visualize the dependence of state transition rates on variables (either parameters or state variables) by plotting a heat map with transitions on one axis and variables on the other. The colour of the heatmap depends on the number of times a variable appears in the expression for the transition rate.

```{r}
m = as.matrix(rs)
ca = corresp(m)
heatmap(m[order(ca$rscore), order(ca$cscore)], 
        Colv = NA, Rowv = NA, scale = "none")
```

I've ordered the axes using correspondence analysis, which is a statistical technique for finding block structure in matrices. Block structure might be relevant because such structure means that we can access parameters in chunks, which could be useful for optimizing on the `C++` side.

### Time Varying Parameters

The model above does not consider time varying parameters, but this will need to be done obviously. My current sense is that for break-point-style parameter changes should be handled when we separate parameters to be optimized from the other parameters (see above). Log-linear-style parameter changes, should probably be handled when we introduce model specification in terms of matrices and vectors (see below).

Now that I've read the MacPan manuscript I feel better about the log-linear model of the transmission rate, $\beta$. A generalization to something beyond the transmission rate is the following.

$$
\log(\theta_i(t)) = \log\left(\theta_i^{(0)}\right) + \mathbf{X}
$$

Each row of $\mathbf{X}$ corresponds to 

### Model Specification in Terms of Matrices and Vectors

The model above treats all parameters and state variables as scalars. But in many cases it is more natural to consider vector- and matrix-valued parameters and states. For example, the contact matrix (`pmat`) that is used by MacPan in models of age-structure.

As far as I can tell models such as age structure _could_ be defined in terms of the model above (because matrix operations are composed of products and sums of products), but it would just require large numbers of tedious entries. There are two ways around this tedium. First, we could add matrix operations to the list of allowable operations in the rate matrix update model. Second, we could create convenience model-specification utilities that would automatically expand to the notation used in the above model. I'm not sure which one I like better, but the first one could take advantage of sparse matrix optimizations from Eigen etc. On the other hand, maybe we can get similar performance with smart index permutations on the R-side?

One issue with the first option is that there is only one kind of scalar multiplication, but several kinds of matrix multiplication -- so we can't just overload the `*` operator (although we probably should as is done in `col_multiply`) and will need to introduce explicit matrix operators (e.g. `%*%`, `%x%`). An issue with Kronecker-type operations (as are used in age-structured models) is that they often involve matrix transposing.

There are some relevant notes on this topic here:  https://github.com/mac-theobio/McMasterPandemic/blob/master/notes/structure.md#general-concerns

## Steve's Notes on Meeting with Ben (2021-08-09)

* very important to be able to specify multiple rates in the model specification machinery -- this is what the regex stuff is doing now
* Additional motivation: figuring out indexing ahead of simulation time is also likely to be an efficiency win.
* 3 components of the model that involve updates of state variables (TODO: verify that the model works for these cases at a minimum):
  * FOI
  * testing -- vec of weights that determines prob of being tested depending on state
  * vaccination -- we know realized vaccination rates, but need to scale to per-capita
* Kronecker products may only be convenient for producing a matrix, rather than efficient at updating it
* most elements of the rate matrix only need to be touched once (update/make rate matrix)
  * this needs to go into the indexing framework
  * by defining in terms of per-capita flows means that most elements can stay fixed
    1. once at beginning of opt run
    2. every time a new set of params are updated
    3. done within the simulation (e.g. state-dependence and time varying parameters
* logs of parameters -- think more about them
* block structure of rate dependence could have a connection with identifiability -- there exists a literature on identifiability of dynamical systems models that boarders on practical
* `TMB::MakeADFun` has a map argument to specify that parameters can be fixed at their starting values or set groups of parameters to be equal -- should use this instead of switching back and forth between the `PARAMETER` and `DATA` macros
